_---
title: "PLAN for Chapter 1. Methodology for fitting ODE epidemic models"
subtitle: "This RMarkdown is a template as initially outlined in the progress report. Each task will be outlined and then followed by a code block to be completed."
output:
html_notebook:
number_sections: true
---


Do the basic comparison between seasonal and non-seasonal model fitting

```{r setup}
library(R.utils)
library(tidyverse)
library(rstan)
library(rstansim) # devtools::install_github("ewan-keith/rstansim")
library(parallel)
library(patchwork)
library(ggridges)
library(pbmcapply)
library(pbapply)
library(memoise)
library(RColorBrewer)
library(ggh4x)
source("../R/constants.R")
source("../R/load_functions.R")

n_cores = parallelly::availableCores() - 1
# n_cores = 9
options(mc.cores = n_cores)
message("Running on ", n_cores, " cores")
rstan_options(auto_write = TRUE)

cd = cachem::cache_disk("memoise")

n_iter = 500 # should be at least 500
n_chains = 1
n_repetitions = 3 # how many times to duplicate each scenario
cores_per_sampler = n_chains # 1 for mclapply, or n_chains if not running lots of scenarios
limit_runs = Inf # set to a finite number for testing, or Inf to run all
timelimit_per_run = 48*hr
n_traces = min(100, n_iter) # limit for plotting sample trajectories
set.seed(0)

start_time = Sys.time()
```

## Model implementation

Generate some data

```{r}
#initial conditions
I_init = 0.01
y0 = c(Il=0, I0=I_init, Sl=0, S0=1-I_init, CumulativeInfections=0)
n_years = 5

# constants for Stan
data_consts = list(y0 = y0,
                   t0 = -100*years,
                   N = 100000, # population size
                   r = 1./60, # r
                   gammal = 1./223, # gammal
                   f = 1./72, # f
                   alpha = 0.21, # alpha
                   beta = 0.66, # beta
                   rho = 0.21, # rho
                   delta = 0 # importation - should ALWAYS be zero for us
)
```

Simulate some dummy data

```{r}
n_times = annual_subdivisions * n_years
data = data_init(data_consts,
                 ts = seq_len(n_times)*years/annual_subdivisions,
                 cases = rep(1, n_times),
                 n_times = n_times,
                 lambda_lower = 0.04,
                 lambda_upper = 0.06)
# params = list(lambda=0.01, phi_inv=0.1, eps=1e-16, kappa=1, phase=130)
params = list(lambda=0.05, phi_inv=0.1, eps=1e-16, kappa=1, phase=130)

model = stan_model("../stan/champagne2022_seasonal_ext_2.stan")

test = compiled_simulate_data(file = model,
                              path = "sim_data",
                              data_name = "seasonal",
                              input_data = data,
                              param_values = params,
                              vars = c("ts", "incidence", "sim_cases"))
sim = my_simulate_data_list(
  file = model,
  path = "sim_data",
  data_name = "seasonal",
  input_data = data,
  param_values = params,
  vars = c("ts", "incidence", "sim_cases")
)

synth_df = with(sim, tibble(
  ts = ts,
  incidence = incidence,
  cases = cases
))

# Edit data with generated values
data_sim = data_init(data_consts,
                     n_times = nrow(synth_df),
                     ts = synth_df$ts,
                     cases = synth_df$cases)

data_sim_agg = aggregate_data(data_sim)
synth_df_agg = with(data_sim_agg, tibble(ts = ts, cases = cases)) 

with(data_sim, tibble(ts = ts/years, cases = cases)) %>%
  ggplot(aes(x=ts, y=cases*12/years)) +
  geom_col() +
  scale_x_continuous(breaks = 0:100)
with(data_sim_agg, tibble(ts = ts/years, cases = cases)) %>%
  ggplot(aes(x=ts, y=cases/years)) +
  geom_col() +
  scale_x_continuous(breaks = 0:100)
```

```{r, eval=F}
fit_seasonal = sampling(stan_model("../stan/champagne2022_seasonal_ext_2.stan"),
                        data = data_sim,
                        iter = n_iter,
                        init = list(params),
                        chains = n_chains)

fit_nonseasonal = sampling(stan_model("../stan/champagne2022.stan"),
                           data = data_sim_agg,
                           iter = n_iter,
                           init = list(params),
                           chains = n_chains)
```

Use Champagne's analytic solution

```{r, eval=F}
mydata = with(data_sim, tibble(
  h = mean(cases * 12 / years) / N * 1,
  # h = mean(cases / years) / N,
  prop_import = 0,
  alpha = alpha,
  beta = beta,
  rho = alpha,
  omega = 1,
  r = 1/60
))
mydata_withR0RC = VivaxModelR::calibrate_vivax_equilibrium(df=mydata, f=1/72, gamma=1/223, r=1/60, return.all = TRUE)
print(mydata_withR0RC)
# fit_analytic =
```

Perform sample importance resampling to turn the analytic solution into a Bayesian solution

```{r, eval=F}
# assume we have at most a 10% error
n_samples_1 = 1000
lambda_domain = seq(params$lambda*0.8, params$lambda*1.1, length.out=n_samples_1)
lambda_prior = dexp(lambda_domain, 1) # effectively uniform, but consistent with rate in stan code

pv_champ_function <- function(t, state, parameters) { # with optional birth death
  with(as.list(c(state, parameters)), {
    FOI = 1*lambda*(Il+I0)
    
    rate.S0_Il = FOI * (1-alpha) # infection without treatment
    rate.S0_Sl = FOI * alpha*(1-beta) # infection with blood stage treatment only
    
    rate.Sl_S0 = gamma + (FOI+f)*alpha*beta # liver clearance + prompt treatment of reinfection/relapse
    rate.Sl_Il = (FOI+f)*(1-alpha) # relapse or infection and no treatment
    
    rate.I0_S0 = FOI*alpha*beta + r # reinfection and full treatment or blood stage recovery
    rate.I0_Il = FOI # reinfection, no treatment
    
    rate.Il_I0 = gamma # liver clearance
    rate.Il_Sl = r # blood stage recovery
    
    dS0 = -rate.S0_Il*S0 - rate.S0_Sl*S0 + rate.Sl_S0*Sl + rate.I0_S0*I0
    dSl = rate.S0_Sl*S0 - rate.Sl_S0*Sl - rate.Sl_Il*Sl + rate.Il_Sl*Il
    dI0 = -rate.I0_S0*I0 + rate.Il_I0*Il - rate.I0_Il*I0
    dIl = rate.S0_Il*S0 + rate.Sl_Il*Sl + rate.I0_Il*I0 - rate.Il_I0*Il - rate.Il_Sl*Il
    
    dCases = (
      FOI * (S0 + Sl) + # force of infection for new blood infections, assume no immunity
        f * Sl # relapses
    ) * alpha * N # detected in the population
    
    list(c(dIl, dI0, dSl, dS0, dCases))
  }) # end with(as.list ...
}

.calculate_likelihood = function(lambda, annual_pop_incidence) {
  pv_state = c(Il = 0.1, I0 = 0, Sl = 0, S0 = 0.9, Cases = 0)
  pv_parameters = with(mydata, list(alpha=alpha, beta=beta, lambda=lambda, delta=0, gamma=1/223, f=1/72, r=1/60, N=data_consts$N))
  
  pv_times = c(0, 10000, 10001)
  pv_sol = deSolve::ode(y = pv_state, times = pv_times, func = pv_champ_function, parms = pv_parameters) %>%
    as_tibble() %>%
    mutate(Cases = Cases - lag(Cases))
  h_total = pv_sol$Cases[length(pv_times)] * 365.25
  likelihood = dpois(round(annual_pop_incidence), h_total)
}
calculate_likelihood = Vectorize(.calculate_likelihood)

observed_annual_cases = round(rev(data_sim_agg$cases)[1])

likelihood = calculate_likelihood(lambda_domain, observed_annual_cases)
plot(lambda_domain, likelihood, 'l')
```

Manually inspect fits

```{r, eval=F}
post_seasonal = rstan::extract(fit_seasonal, c("lambda", "eps", "kappa", "phase")) %>%
  as_tibble() %>%
  pivot_longer(everything())

ggplot(post_seasonal, aes(x=value, color=name, fill=name)) +
  geom_density() +
  facet_wrap(vars(name), scales="free")

incidence_seasonal = extract_incidence(fit_seasonal, data_sim$ts)
ggplot(incidence_seasonal, aes(x=ts, y=incidence, group=trace)) +
  geom_line(alpha = 0.05) +
  geom_point(data = synth_df, aes(y=cases, group=NULL)) +
  coord_cartesian(ylim = c(0, NA))

post_nonseasonal = rstan::extract(fit_nonseasonal, c("lambda")) %>%
  as_tibble() %>%
  pivot_longer(everything())

ggplot(post_nonseasonal, aes(x=value, color=name, fill=name)) +
  geom_density() +
  facet_wrap(vars(name), scales="free")

incidence_nonseasonal = extract_incidence(fit_nonseasonal, data_sim_agg$ts)
ggplot(incidence_nonseasonal, aes(x=ts, y=incidence, group=trace)) +
  geom_line(alpha = 0.05) +
  geom_point(data = synth_df_agg, aes(y=cases, group=NULL)) +
  coord_cartesian(ylim = c(0, NA))
```

# Can we make a variety of scenarios?
Define a function to simulate cases for a scenario

```{r}
model = stan_model("../stan/champagne2022_seasonal_ext_2.stan")
simulate_cases = function(data_consts, lambda, phi_inv, eps, kappa, phase) {
  n_times = annual_subdivisions * n_years
  data = data_init(data_consts,
                   cases = rep(1, n_times),
                   ts = seq_len(n_times) * years / annual_subdivisions,
                   n_times = n_times)
  params = list(lambda=lambda, phi_inv=phi_inv, eps=eps, kappa=kappa, phase=phase)
  synth_df = my_simulate_data(
    file = model,
    path = "sim_data",
    data_name = "seasonal",
    input_data = data,
    param_values = params,
    vars = c("ts", "incidence", "sim_cases")
  )
}
```

Create scenarios

```{r}
# Define the scenarios
range_lambda = seq(0.010, 0.011, length.out=3)
range_eps = seq(1e-12, 1-1e-12, length.out=3)
scenarios = expand_grid(
  lambda = range_lambda,
  phi_inv = 0.1,
  eps = range_eps,
  kappa = 1,
  phase = years/2,
  rep = seq_len(n_repetitions)
) %>%
  mutate(ID = row_number(), .before=0)

# Simulate cases for each scenario
cases_scenarios = lapply(seq_len(nrow(scenarios)), function(i) {
  lambda = scenarios$lambda[i]
  phi_inv = scenarios$phi_inv[i]
  eps = scenarios$eps[i]
  kappa = scenarios$kappa[i]
  phase = scenarios$phase[i]
  simulate_cases(data_consts, lambda, phi_inv, eps, kappa, phase)
})

# Attach simulation results to the scenarios dataframe
scenarios$ts = lapply(cases_scenarios, function(x) {x$ts})
scenarios$cases = lapply(cases_scenarios, function(x) {x$cases})

# Duplicate for each method
scenarios_long = scenarios %>%
  cross_join(tibble(method = c("../stan/champagne2022_seasonal_ext_2.stan",
                               "../stan/champagne2022.stan")))
# "../stan/champagne2022_poisson.stan")))
# Replace data with aggregate data depending on the method
for (i in seq_len(nrow(scenarios_long))) {
  if (!scenarios_long$method[i] %>% str_detect("seasonal")) {
    scenarios_long$ts[[i]] = aggregate_time(scenarios_long$ts[[i]])
    scenarios_long$cases[[i]] = aggregate_cases(scenarios_long$cases[[i]])
  }
}
```

```{r eval=F}
# Add champagne solution
lambda_champagne = function(ts, cases, data) {
  dt = ts[2] - ts[1]
  mydata = with(data, tibble(
    h = mean(cases / dt) / N,
    prop_import = 0,
    alpha = alpha,
    beta = beta,
    rho = rho,
    omega = 1,
    r = r
  ))
  
  mydata_withR0RC = suppressMessages(VivaxModelR::calibrate_vivax_equilibrium(df=mydata, f=data$f, gamma=data$gamma, r=data$r, return.all = TRUE))
  
  return(mydata_withR0RC$lambda)
}

scenarios_long$lambda_champagne = NA_real_
for (i in seq_len(nrow(scenarios_long))) {
  cases = scenarios_long$cases[[i]]
  ts = scenarios_long$ts[[i]]
  data = data_consts
  
  dt = ts[2] - ts[1]
  print(dt)
  mydata = with(data, tibble(
    h = mean(cases / dt) / N,
    prop_import = 0,
    alpha = alpha,
    beta = beta,
    rho = rho,
    omega = 1,
    r = r
  ))
  
  mydata_withR0RC = suppressMessages(VivaxModelR::calibrate_vivax_equilibrium(df=mydata, f=data$f, gamma=data$gamma, r=data$r, return.all = TRUE))
  
  lambda_champagne = mydata_withR0RC$lambda
  
  scenarios_long$lambda_champagne[i] = lambda_champagne
}

scenarios_long$lambda_champagne

ggplot(scenarios_long, aes(x=lambda, y=lambda_champagne, color=method)) +
  geom_point(alpha = 0.8) +
  geom_abline(slope=1, intercept=0)

scenarios_long %>%
  select(lambda, lambda_champagne, method)
```

What do our synthetic data look like?

```{r}
scenarios_long %>%
  mutate(method = case_when(method %>% str_detect("seasonal") ~ "Seasonal",
                            TRUE ~ "Nonseasonal")) %>%
  # distinct(scenario_ID, .keep_all=T) %>%
  unnest(cols = c("ts", "cases")) %>%
  mutate(cases = ifelse(method == "Nonseasonal", cases/annual_subdivisions, cases)) %>%
  ggplot(aes(x = ts/years, y = cases, color=method, group=interaction(method, rep))) +
  geom_line(alpha = 0.6) +
  scale_x_continuous(breaks = 0:n_years) +
  scale_y_log10(labels = label_auto2) +
  facet_grid(rows = vars(eps),
             cols = vars(lambda),
             labeller = plot_labeller) +
  labs(title = "Input data for each scenario",
       subtitle = "Seasonal scenario in cases per month; nonseasonal scenario per year",
       x = "Year",
       y = "Monthly cases",
       color = "Method")
ggsave("../plots/basic_data.png", width=8, height=4)
```

Define a function to perform the model fit, then apply it to each scenario

```{r}
# Define function that takes simulated data
.my_sampling = function(method, ts, cases, true_lambda=NULL) {
  data = data_init(data_consts,
                   n_times = length(ts),
                   ts = ts,
                   cases = cases,
                   lambda_lower = 0.009,
                   lambda_upper = 0.013)
  if (method %>% str_detect("seasonal")) {
    result = sampling(stan_model(method),
                      data = data,
                      iter = n_iter,
                      init = list(list(lambda = true_lambda)),
                      chains = 1,
                      pars = c("lambda", "phi_inv", "eps", "kappa", "phase", "incidence"))
  } else {
    result = sampling(stan_model(method),
                      data = data,
                      iter = n_iter,
                      init = list(list(lambda = true_lambda)),
                      chains = 1,
                      pars = c("lambda", "phi_inv", "incidence"))
  }
  result
}
# my_sampling = memoise::memoise(.my_sampling, cache = cd)
# memoise::forget(my_sampling)
my_sampling = .my_sampling
# my_sampling = .my_sampling

# Apply function to simulated data
tictoc::tic("Running all scenarios")
fit_ix = function(i) {
  result = with_timeout(
    my_sampling(scenarios_long$method[i], scenarios_long$ts[[i]], scenarios_long$cases[[i]]),
    cpu = timelimit_per_run)
  
  # if (is.null(result)) {
  #   memoise::drop_cache(my_sampling)(scenarios_long$method[i], scenarios_long$ts[[i]], scenarios_long$cases[[i]])
  # } else if (!"stanfit" %in% class(result)) {
  #   memoise::drop_cache(my_sampling)(scenarios_long$method[i], scenarios_long$ts[[i]], scenarios_long$cases[[i]])
  # } else if (result@mode == 2) {
  #   memoise::drop_cache(my_sampling)(scenarios_long$method[i], scenarios_long$ts[[i]], scenarios_long$cases[[i]])
  # }
  
  return(result)
}
ix_nonseasonal = which(!str_detect(scenarios_long$method, "easonal"))
ix_seasonal = which(str_detect(scenarios_long$method, "easonal"))
fits_nonseasonal = pbmclapply(ix_nonseasonal, fit_ix)
fits_seasonal = pbmclapply(ix_seasonal, fit_ix)
# assign seasonal and nonseasonal fits back into the original list
setClass("Nullsim", representation=representation(mode = "numeric"))
fits_scenarios = as.list(rep(NA, nrow(scenarios_long)))
for (i in seq_along(ix_seasonal)) {
  j = ix_seasonal[i]
  fits_scenarios[[j]] = fits_seasonal[[i]]
  # fits_scenarios[[j]] = new("Nullsim", mode=2)
}
for (i in seq_along(ix_nonseasonal)) {
  j = ix_nonseasonal[i]
  fits_scenarios[[j]] = fits_nonseasonal[[i]]
}
# Add null sims instead of nulls or errors
for (i in seq_along(fits_scenarios)) {
  fit = fits_scenarios[[i]]
  if (is.null(fit) | "try-error" %in% class(fit)) {
    fits_scenarios[[i]] = new("Nullsim", mode=2)
  }
}
tictoc::toc()
```

How many didn't time out?

```{r}
sapply(fits_scenarios, is.null)
sapply(fits_scenarios, function(x) {x@mode})

# Redo any that didn't work
ix_redo = which(sapply(fits_scenarios, is.null) | sapply(fits_scenarios, function(x) {x@mode == 2}))
# fits_scenarios[[ix_redo]] = pbmclapply(ix_redo, fit_ix)
if (length(ix_redo) == 1) {
  fits_scenarios[[ix_redo]] = lapply(list(ix_redo), fit_ix)
} else if (length(ix_redo) > 1) {
  fits_scenarios[[ix_redo]] = pbmclapply(ix_redo, fit_ix)
}

# Add null sims instead of nulls or errors
for (i in seq_along(fits_scenarios)) {
  fit = fits_scenarios[[i]]
  if (is.null(fit) | "try-error" %in% class(fit)) {
    fits_scenarios[[i]] = new("Nullsim", mode=2)
  }
}
```

```{r}
# Extract parameters into dataframe
estimates_scenarios = lapply(seq_len(length(fits_scenarios)), function(i) {
  fit = fits_scenarios[[i]]
  method = scenarios_long$method[i]
  if (fit@mode == 2) {
    return(NULL)
  }
  #print(fit@mode)
  if (scenarios_long$method[i] %>% str_detect("seasonal")) {
    posterior = rstan::extract(fit, c("lambda", "phi_inv", "eps", "kappa", "phase")) %>%
      as_tibble()
  } else {
    posterior = rstan::extract(fit, c("lambda", "phi_inv")) %>%
      as_tibble() %>%
      mutate(eps = NA, kappa = NA, phase = NA)
  }
  return(posterior)
})


# Attach re-estimated parameters
scenarios_long$est_lambda = lapply(estimates_scenarios, function(x) {x$lambda})
scenarios_long$est_phi = lapply(estimates_scenarios, function(x) {1/x$phi_inv})
scenarios_long$phi = 1/scenarios_long$phi_inv
scenarios_long$est_eps = lapply(estimates_scenarios, function(x) {x$eps})
scenarios_long$est_kappa = lapply(estimates_scenarios, function(x) {x$kappa})
scenarios_long$est_phase = lapply(estimates_scenarios, function(x) {x$phase})

# Attach posterior predictions
scenarios_long$est_incidence = lapply(seq_len(nrow(scenarios_long)), function(i) {
  x = fits_scenarios[[i]]
  if (x@mode == 2) {
    return(NULL)
  }
  scenario = scenarios_long[i,]
  method = scenario$method
  data_sim = data_init(data_consts, n_times = length(scenario$ts[[1]]), ts = scenario$ts[[1]], cases = scenario$cases[[1]])
  incidence = rstan::extract(x, "incidence")[[1]]
  if (n_traces < dim(incidence)[1]) {
    sample_ix = seq_len(dim(incidence)[1])
  } else {
    sample_ix = sample(seq_len(dim(incidence)[1]), n_traces, replace=T)
  }
  
  incidence_sample_nonseasonal = as_tibble(t(incidence[sample_ix,])) %>%
    mutate(j = row_number()) %>%
    pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
    drop_na(j) %>%
    group_by(j) %>%
    mutate(ts = data_sim$ts[j],
           lower = quantile(incidence, 0.025, na.rm=T),
           upper = quantile(incidence, 0.975, na.rm=T),
           legend = "95% prediction interval",
           rep = scenario$rep)
})

# scenarios_long = scenarios_long %>%
#   filter(lambda > 0.01)
```

Diagnostic plots

Firstly, inspect the seasonal model.

```{r}
plot_data = scenarios_long %>%
  filter(method %>% str_detect("seasonal")) %>%
  select(-ts, -cases, -est_incidence) %>%
  pivot_longer(matches("^est")) %>%
  unnest(value) %>%
  drop_na(value) %>%
  mutate(name = name %>% str_remove("^est_"),
         method = ifelse(method %>% str_detect("seasonal"), "Seasonal", "Nonseasonal")) %>%
  group_by(ID, name) %>%
  mutate(lq = quantile(value, 0.01),
         uq = quantile(value, 0.99)) %>%
  filter(value < uq) %>%
  ungroup()

true_values = scenarios_long %>%
  distinct(lambda, phi, eps, kappa, phase) %>%
  mutate(lambda2 = lambda,
         eps2 = eps) %>%
  pivot_longer(-c(lambda2, eps2)) %>%
  dplyr::rename(lambda = lambda2,
                eps = eps2)

plot_seasonal_posterior = ggplot(plot_data, aes(x = value, y = as.factor(lambda), color = as.factor(rep), fill = as.factor(rep))) +
  geom_density_ridges(alpha = 0.5) +
  geom_point(data = true_values, aes(color=NULL, fill = NULL, group=NULL), shape=1) +
  facet_grid(rows = vars(eps), cols = vars(name), scales="free", labeller = labeller(name = ~ make_greek(as.character(.)),
                                                                                     eps = ~ paste(make_greek("eps"), "=", round(as.numeric(.), 5)),
                                                                                     .multi_line=F)) +
  
  scale_color_brewer(palette="Set2") +
  scale_fill_brewer(palette="Set2") +
  labs(title = "Posterior parameter densities of the seasonal model",
       subtitle ="True values denoted by circles",
       x = "Value",
       y = paste("Scenario", make_greek("lambda")),
       color = paste("Scenario", make_greek("lambda")),
       fill = paste("Scenario", make_greek("lambda"))) +
  theme(legend.position = "none")
plot_seasonal_posterior

ggsave("../plots/basic_seasonal_posterior.png", width=8, height=4)
```

Do the same for the nonseasonal model

```{r}
plot_data = scenarios_long %>%
  filter(!method %>% str_detect("seasonal")) %>%
  select(-ts, -cases, -est_incidence) %>%
  pivot_longer(matches("^est")) %>%
  unnest(value) %>%
  drop_na(value) %>%
  mutate(name = name %>% str_remove("^est_"),
         method = ifelse(method %>% str_detect("seasonal"), "Seasonal", "Nonseasonal")) %>%
  group_by(name) %>%
  mutate(lq = quantile(value, 0.01),
         uq = quantile(value, 0.99)) %>%
  filter(value < uq) %>%
  ungroup()

true_values = scenarios_long %>%
  distinct(lambda, phi, eps, kappa, phase) %>%
  mutate(lambda2 = lambda,
         eps2 = eps) %>%
  pivot_longer(-c(lambda2, eps2)) %>%
  dplyr::rename(lambda = lambda2,
                eps = eps2) %>%
  filter(name %in% plot_data$name)

plot_nonseasonal_posterior = ggplot(plot_data, aes(x = value, y = as.factor(lambda), color = as.factor(rep), fill = as.factor(rep))) +
  geom_density_ridges(alpha = 0.5) +
  geom_point(data = true_values, aes(color=NULL, fill = NULL, group=NULL), shape=1) +
  ggh4x::facet_grid2(rows = vars(eps), cols = vars(name), scales="free", labeller = labeller(name = ~ make_greek(as.character(.)),
                                                                                             eps = ~ paste(make_greek("eps"), "=", round(as.numeric(.), 5)),
                                                                                             .multi_line=F),
                     independent=TRUE) +
  scale_color_brewer(palette="Set2") +
  scale_fill_brewer(palette="Set2") +
  labs(title = "Posterior parameter densities of the nonseasonal model",
       subtitle ="True values denoted by circles",
       x = "Value",
       y = paste("Scenario", make_greek("lambda")),
       color = paste("Scenario", make_greek("lambda")),
       fill = paste("Scenario", make_greek("lambda"))) +
  theme(legend.position = "none")
plot_nonseasonal_posterior

ggsave("../plots/basic_nonseasonal_posterior.png", width=8, height=4)

(plot_seasonal_posterior + labs(title=NULL, subtitle="Seasonal model")) / (plot_nonseasonal_posterior + labs(title=NULL, subtitle="Nonseasonal model")) + plot_annotation(title = "Posterior parameter densities colored by repetition", tag_levels="A")
ggsave("../plots/basic_posterior.png", width=8, height=8)
```

Then compare some parameters between seasonal and nonseasonal models

```{r}
plot_data = scenarios_long %>%
  select(-ts, -cases, -est_incidence) %>%
  pivot_longer(matches("^est")) %>%
  unnest(value) %>%
  drop_na(value) %>%
  mutate(name = name %>% str_remove("^est_"),
         method = ifelse(method %>% str_detect("seasonal"), "Seasonal", "Nonseasonal")) %>%
  # filter(method == "Seasonal") %>%
  group_by(name, method) %>%
  mutate(lq = quantile(value, 0.01),
         uq = quantile(value, 0.99)) %>%
  filter(value >= lq,
         value <= uq) %>%
  ungroup()

true_values = scenarios_long %>%
  distinct(lambda, phi, eps, kappa, phase) %>%
  mutate(lambda2 = as.factor(lambda),
         eps2 = as.factor(eps)) %>%
  pivot_longer(-c(lambda2, eps2)) %>%
  dplyr::rename(lambda = lambda2,
                eps = eps2)

posterior_ranges = tribble(
  ~name, ~min, ~max,
  "lambda", 0, 0.015,
  "phi", 0, 20,
  "eps", 0, 1,
  "kappa", 0, 10,
  "phase", 0, years
)
posterior_ranges_data = posterior_ranges %>%
  pivot_longer(-name, names_to="bound")

plot_data = plot_data %>%
  inner_join(posterior_ranges, by="name") %>%
  filter(value >= min, value <= max)

ggplot(plot_data, aes(x = value, y = as.factor(lambda), color = method, fill = method, group = interaction(lambda, rep, method))) +
  geom_density_ridges(alpha = 0.25) +
  geom_point(data = posterior_ranges_data, aes(y = as.factor(0.01), color = NULL, fill = NULL, group = NULL), alpha=0) +
  geom_point(data = true_values, aes(color=NULL, fill = NULL, group=NULL), shape=1) +
  facet_wrap(vars(eps, make_greek(name)), scales="free", nrow=3, labeller = labeller(eps = ~ paste(make_greek("eps"), "=", round(as.numeric(.), 5)),
                                                                                     .multi_line=F)) +
  theme(legend.position = "right") +
  labs(title = "Synthetic parameter estimation comparison over repeated scenarios",
       x = "Value", y = NULL, color = "Method", fill = "Method")
ggsave("../plots/basic_posterior_comparison.png", width=8, height=4)
```

Common parameters only

```{r}
plot_data_2 = plot_data %>%
  filter(name == "lambda") %>%
  mutate(eps = as.numeric(as.character(eps)))
true_values_2 = true_values %>%
  filter(name == "lambda") %>%
  mutate(eps = as.numeric(as.character(eps))) %>%
  left_join(tibble(method = c("Seasonal", "Nonseasonal")), by=character())

xbreaks = unique(true_values_2$value)
dx = xbreaks[2] - xbreaks[1]
xbreaks = c(min(xbreaks)-dx, xbreaks, max(xbreaks)+dx)
xrange = c(min(xbreaks), max(xbreaks))

p1 = ggplot(plot_data_2, aes(x = value, y = 1, color = method, fill = method, group = interaction(lambda, rep, method))) +
  geom_vline(data=true_values_2, aes(xintercept = value), linetype="dashed") +
  geom_density_ridges(alpha = 0.25) +
  # geom_density(aes(y=NULL)) +
  scale_x_continuous(breaks = xbreaks, minor_breaks=NULL) +
  expand_limits(x=xrange) +
  # coord_cartesian(xlim = c(min(xbreaks), max(xbreaks))) +
  # geom_point(data = true_values_2, aes(y = method, color=NULL, fill = NULL, group=NULL), shape=1) +
  facet_grid(rows = vars(eps), cols = vars(lambda), labeller = plot_labeller, scales = "free_y") +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(subtitle = paste("Estimates of", make_greek("lambda")),
       x = "Value", y = NULL, color = "Method", fill = "Method")

plot_data_2 = plot_data %>%
  filter(name == "phi") %>%
  mutate(eps = as.numeric(as.character(eps)))
true_values_2 = true_values %>%
  filter(name == "phi") %>%
  mutate(eps = as.numeric(as.character(eps))) %>%
  left_join(tibble(method = c("Seasonal", "Nonseasonal")), by=character())

p2 = ggplot(plot_data_2, aes(x = value, y = 1, color = method, fill = method, group = interaction(lambda, rep, method))) +
  geom_vline(data=true_values_2, aes(xintercept = value), linetype="dashed") +
  geom_density_ridges(alpha = 0.25) +
  # geom_density(aes(y=NULL)) +
  # scale_x_continuous(breaks = unique(true_values_2$value), minor_breaks=NULL) +
  # geom_point(data = true_values_2, aes(y = method, color=NULL, fill = NULL, group=NULL), shape=1) +
  facet_grid(rows = vars(eps), cols = vars(lambda), labeller = plot_labeller, scales = "free_y") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(subtitle = paste("Estimates of", make_greek("phi")),
       x = "Value", y = NULL, color = "Method", fill = "Method")

p1 + p2 + plot_layout(guides="collect") + plot_annotation(title = "Parameter estimates over repeated scenarios")

ggsave("../plots/basic_method_comparison.png", width=8, height=4)
```

Plot incidence fits

```{r}
incidence_join = lapply(seq_len(nrow(scenarios_long)), function(i) {
  if (is.null(scenarios_long$est_incidence[[i]])){
    return(NULL)
  }
  est_incidence = scenarios_long$est_incidence[[i]] %>%
    mutate(ID = scenarios_long$ID[i],
           method = scenarios_long$method[i],
           lambda = scenarios_long$lambda[i],
           eps = scenarios_long$eps[i],
           rep = scenarios_long$rep[i])
}) %>%
  bind_rows() %>%
  ungroup() %>%
  select(-lower, -upper, -legend)

if (n_traces < length(unique(incidence_join$trace))) {
  trace_sample = sample(unique(incidence_join$trace), n_traces)
  incidence_join = incidence_join %>%
    filter(trace %in% trace_sample)
}

# For each scenario, find a rep where nothing failed
first_successful_scenario = incidence_join %>%
  group_by(lambda, eps, method) %>%
  slice(1) %>%
  select(lambda, eps, method, rep) %>%
  ungroup()
incidence_join = incidence_join %>%
  inner_join(first_successful_scenario, by=c("rep", "method", "lambda", "eps")) %>%
  mutate(method = ifelse(method %>% str_detect("seasonal"), "Seasonal", "Nonseasonal")) %>%
  mutate(incidence = ifelse(method == "Nonseasonal", incidence/12, incidence))


scenarios_overlay = scenarios %>%
  unnest(c(ts, cases)) %>%
  inner_join(first_successful_scenario, by=c("rep", "lambda", "eps"))

ggplot(incidence_join, aes(x = ts/years, y = incidence, color=method, group = interaction(method, rep, trace))) +
  geom_point(data=scenarios_overlay, aes(y=cases, color=NULL, group=NULL), alpha=0.25, show.legend=F, size=1) +
  geom_line(alpha = 0.5) +
  facet_grid(rows = vars(eps), cols = vars(lambda), labeller = plot_labeller, scales = "free_y") +
  # facet_grid2(rows = vars(eps), cols = vars(lambda), scales="free_y", independent="y") +
  scale_x_continuous(breaks = 0:n_years) +
  scale_y_log10(labels = label_auto2, limits=c(1, NA)) +
  labs(title = "Mean estimated incidence rates for the first scenario repetition",
       subtitle = "Nonseasonal (annual) results scaled to per-month",
       x = "Year",
       y = "Incidence", color = "Method") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))

ggsave("../plots/basic_modelled_incidence.png", width=8, height=4)
```

Re-simulate data with the true seasonality

```{r}
valid_scenarios_long = scenarios_long %>%
  inner_join(first_successful_scenario)
resimulated = lapply(seq_len(nrow(valid_scenarios_long)), function(i) {
  # if (is.null(valid_scenarios_long$est_incidence[[i]])){
  #   return(NULL)
  # }
  lapply(seq_len(n_traces), function(j) {
    if (is.null(valid_scenarios_long$est_lambda[[i]])) {
      return(NULL)
    }
    lambda = sample(valid_scenarios_long$est_lambda[[i]], 1)
    phi_inv = 1e-10 # valid_scenarios_long$phi_inv[i] # Doesn't matter
    eps = valid_scenarios_long$eps[i]
    kappa = valid_scenarios_long$kappa[i]
    phase = valid_scenarios_long$phase[i]
    x = simulate_cases(data_consts, lambda, phi_inv, eps, kappa, phase)
    if (is.data.frame(x)) {
      result = x %>%
        mutate(lambda = valid_scenarios_long$lambda[i],
               eps = eps,
               method = valid_scenarios_long$method[i],
               trace = as.character(j))
      return(result)
    } else {
      return(NULL)
    }
    
  }) %>%
    bind_rows()
})
resimulated_2 = resimulated %>%
  bind_rows() %>%
  mutate(method = ifelse(method %>% str_detect("seasonal"), "Seasonal", "Nonseasonal"),
         trace = as.character(trace))# %>%
# filter(method == "Nonseasonal")

ggplot(resimulated_2, aes(x = ts/years, y = incidence, color=method, group = interaction(method, trace))) +
  geom_point(data=scenarios_overlay, aes(y=cases, color=NULL, group=NULL), alpha=0.25, show.legend=F, size=1) +
  geom_line(alpha = 0.1) +
  facet_grid(rows = vars(eps), cols = vars(lambda), labeller = plot_labeller) +
  scale_x_continuous(breaks = 0:n_years) +
  scale_y_log10(labels = label_auto2, limits=c(1, NA)) +
  labs(title = "Re-simulated incidence with true seasonality for the first scenario repetition",
       x = "Year",
       y = "Incidence", color = "Method") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))

ggsave("../plots/basic_resimulation.png", width=8, height=4)
```

Compare the nonseasonal resimulation against the seasonal fit

```{r}
incidence_comparison = incidence_join %>%
  mutate(method = paste(method, "model")) %>%
  bind_rows(resimulated_2 %>%
              filter(method == "Nonseasonal") %>%
              mutate(method = "Nonseasonal resimulation"))

scenarios_overlay_2 = scenarios_overlay %>%
  mutate(ts = ts %% years + years)

ggplot(incidence_comparison, aes(x = ts/years*annual_subdivisions, y = incidence, color=method, group = interaction(method, trace))) +
  geom_line(alpha = 0.2) +
  geom_point(data=scenarios_overlay_2, aes(y=cases, color=NULL, group=NULL), alpha=0.2, show.legend=F, size=0.5) +
  ggh4x::facet_grid2(rows=vars(eps), cols=vars(lambda), labeller=plot_labeller, scales="free_y", independent="y") +
  scale_x_continuous(breaks = annual_subdivisions + seq_len(annual_subdivisions), labels=sapply(month.abb, function(x) {substr(x,0,1)})) +
  coord_cartesian(xlim = c(annual_subdivisions+1, annual_subdivisions*2)) +
  # scale_y_log10(labels = label_auto2, limits=c(NA, NA)) +
  labs(title = "Prediction comparison for the first repetition",
       subtitle = "Calibration data condensed by month",
       x = "Year",
       y = "Incidence", color = "Method") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))

ggsave("../plots/basic_resimulation_comparison.png", width=8, height=4)
```

```{r}
end_time = Sys.time()
print(end_time)
print(end_time - start_time)
workspace_filename = paste0("workspaces/Chapter_01_basic_", Sys.Date(), ".RData")
save.image(workspace_filename)
if (hour(Sys.time()) >= 9) {beepr::beep()}
```