---
title: "PLAN for Chapter 2. Simulation-estimation study for the temperate model"
subtitle: "This RMarkdown is a template as initially outlined in the progress report. Each task will be outlined and then followed by a code block to be completed."
output:
html_notebook:
number_sections: true
---

```{r setup}
library(R.utils)
library(tidyverse)
library(rstan)
library(parallel)
library(patchwork)
library(pbmcapply)
library(pbapply)
library(memoise)
library(RColorBrewer)
source("../R/constants.R")
source("../R/load_functions.R")

# n_cores = parallelly::availableCores()
rstan_options(auto_write = TRUE)

n_iter = 1000 # should be at least 500
n_chains = 4
n_cores = n_chains
n_traces = 200 # limit for plotting sample trajectories
options(mc.cores = n_cores)
message("Running on ", n_cores, " cores")

# Compile model before we reach any sampling
model = stan_model("../stan/temperate_11.stan")
my_state_init = state_init_3
my_extract_summary = extract_summary_2
start_time = Sys.time()
```
## Parameter recovery on real data

We test whether parameter recovery works on real datasets. We will find data from a variety of settings (e.g., transmission levels, remoteness, strains) to demonstrate generalisability.

Data:

- Hainan data (tropical China) but this has been difficult to acquire.
We will not be investigating Chinese Yunnan (southern mountainous) or Henan (central temperate) data which we do have because the Yunnan strain is not known to be tropical, and the Henan data is temperate (our current temperate model does not align with this data convincingly).
- Brazilian 'integrated data set', available per county or municipality and very detailed.

```{r}
china_selections = tribble(
  ~Region, ~min, ~max,
  "Dengzhou", "2004-01-01", "2009-01-01",
  "Guantang", "1977-01-01", "1982-01-01",
  "Huangchuan", NA, NA,
  "Xiayi", NA, NA
) %>%
  mutate(min = as.Date(min),
         max = as.Date(max))

library(MalariaData)
regions = c("Xiayi", "Guantang", "Dengzhou", "Huangchuan") %>% setNames({.})
china_data_all = lapply(regions, function(x) {load_region(x, species = "Vivax", source="Local")}) %>%
  bind_rows(.id = "Region") %>%
  select(Region, Date, Cases) %>%
  left_join(china_selections, by="Region") %>%
  mutate(Date = as.Date(paste(year(Date), month(Date), "01", sep="-")) + months(1) - days(1)) %>% # Align to last day of the month
  mutate(include = ifelse(Date >= min & Date < max, "grey20", "grey"),
         include = replace_na(include, "grey"))

china_data = china_data_all %>%
  filter(include == "grey20") %>%
  select(-include)
```

What if we fit it with no clinical immunity?

```{r}
selected_regions = c("Dengzhou", "Guantang") %>% setNames({.})
samp_results = list()
data_regions = lapply(selected_regions, function(region) {
  ts_region = china_data %>%
    filter(Region == region)
  first_year = min(year(ts_region$Date))
  ts_region = ts_region %>%
    mutate(ts = as.numeric(Date - as.Date(paste0(first_year, "-01-01"))))
  data_region = list(
    t0 = -30*years,
    n_times = nrow(ts_region),
    ts = ts_region$ts,
    cases = ts_region$Cases,
    relapse_clinical_immunity = 0,
    gamma_d = 1/434.,
    gamma_l = 1/223,
    # delta = 1/162,
    delta = 1/220,
    phi = 0,
    f = 1/72,
    r = 1/60,
    p_long = 1,
    p_silent = 0.34, # includes a proportion of delayed primary infections
    N = 10000,
    population_size = 10000,
    n_dormant = 11,
    eps = 0,
    # Prior shapes
    ## Beta(1, 1) is uniform
    relapse_clinical_immunity_shape1 = 1,
    relapse_clinical_immunity_shape2 = 50,
    p_silent_shape1 = 15,
    p_silent_shape2 = 30
  )
  data_region$y0 = my_state_init(data_region, I0=0.01)
  return(data_region)
})

for (selected_region in selected_regions) {
  data_region = data_regions[[selected_region]]
  
  tictoc::tic()
  samp = sampling(model,
                  data = data_region,
                  iter = n_iter,
                  chains = n_chains,
                  cores = n_chains,
                  pars = c("y", "y_extended"),
                  include = F)
  tictoc::toc()
  samp_results[[selected_region]] = samp
}
```

```{r}
# inspect parameter posteriors
parameters = c("lambda", "phi", "alpha", "beta", "eps", "kappa", "phase", "relapse_clinical_immunity", "p_long", "p_silent")

posterior_seasonal = lapply(samp_results, function(samp) {
  posterior_seasonal = rstan::extract(samp, parameters[parameters %in% names(samp)]) %>%
    as_tibble() %>%
    pivot_longer(everything(), names_to = "parameter", values_to = "value")
}) %>%
  bind_rows(.id = "Region")

var_plot = posterior_seasonal %>%
  ggplot(aes(x = value, fill=Region, color=Region)) +
  geom_density(alpha=0.5) +
  # scale_fill_manual(values = param_colors, drop=F) +
  # scale_color_manual(values = param_colors, drop=F) +
  coord_cartesian(xlim = c(0, NA)) +
  facet_wrap(vars(parameter), scales="free", labeller=plot_labeller_novar) +
  labs(subtitle = "Parameter estimates")
# var_plot
# 
# filename = paste0("../plots/china_variables_plausible.png")
# ggsave(filename, width=8, height=4)

# Plot credible intervals
traces = lapply(samp_results, function(samp) {
  incidence = rstan::extract(samp, "incidence")[[1]]
  cases = rstan::extract(samp, "sim_cases")[[1]]
  incidence_sample_temperate = as_tibble(t(incidence)) %>%
    mutate(j = row_number()) %>%
    pivot_longer(-j, names_to = "trace", values_to = "value")
  cases_sample_temperate = as_tibble(t(cases)) %>%
    mutate(j = row_number()) %>%
    pivot_longer(-j, names_to = "trace", values_to = "value")
  
  bind_rows(`Post. pred. cases` = cases_sample_temperate,
            `Mean incidence` = incidence_sample_temperate,
            .id = "metric") %>%
    mutate(metric = fct_inorder(metric)) %>%
    group_by(j, metric) %>%
    summarise(ts = data_region$ts[j],
              lower = quantile(value, 0.025, na.rm=T),
              upper = quantile(value, 0.975, na.rm=T),
              legend = "95% prediction interval") %>%
    ungroup()
}) %>%
  bind_rows(.id = "Region")

plot_original_data = lapply(data_regions, function(x) {
  tibble(ts = x$ts,
         cases = x$cases)
}) %>%
  bind_rows(.id = "Region")

epi_plot = ggplot(mapping = aes(x=ts/years)) +
  geom_ribbon(data = traces,
              aes(ymin=lower, ymax=upper, fill=metric),
              alpha = 0.5) +
  geom_point(data = plot_original_data,
             aes(y = cases, group = NULL, color="Case observations")) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_fill_manual("", 
                    breaks = c("Post. pred. cases", "Mean incidence", "Incidence data"),
                    values = c("steelblue", "tomato", "black")) +
  scale_colour_manual("", 
                      breaks = c("Post. pred. cases", "Mean incidence", "Case observations"),
                      values = c("steelblue", "tomato", "black")) +
  facet_wrap(vars(Region), scales="free_y") +
  labs(x = "Year",
       y = "Monthly incidence")
# epi_plot
# 
# filename = paste0("../plots/china_incidence_plausible.png")
# ggsave(filename, width=8, height=4)

epi_plot / var_plot + plot_annotation(tag_levels="A")
filename = paste0("../plots/china_posterior_plausible.png")
ggsave(filename, width=8, height=8)
```

Save workspace

```{r}
end_time = Sys.time()
print(end_time)
print(end_time - start_time)
workspace_filename = paste0("workspaces/Chapter_02_china_plausible_", Sys.Date(), ".RData")
save.image(workspace_filename)
beepr::beep()
```

Plot traces

```{r}
region_summaries = lapply(names(samp_results) %>% setNames({.}), function(region) {
  my_sample = samp_results[[region]]
  my_data = data_regions[[region]]
  my_summary = my_extract_summary(my_data, my_sample)
}) %>%
  bind_rows(.id = "Region") %>%
  pivot_longer(c(Infectious, Dormant, Latent, Susceptible)) %>%
  select(Region, time, name, value)

region_summaries %>%
  ggplot(aes(x = time / years, y = value, color = name)) +
  geom_line() +
  facet_grid(cols = vars(Region)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Year", y = "Proportion", color = "Compartment")
```
