_---
title: "PLAN for Chapter 1. Methodology for fitting ODE epidemic models"
subtitle: "This RMarkdown is a template as initially outlined in the progress report. Each task will be outlined and then followed by a code block to be completed."
output:
html_notebook:
number_sections: true
---

This is to test a model with variable numbers of compartments.

```{r setup}
library(R.utils)
library(tidyverse)
library(rstan)
library(rstansim) # devtools::install_github("ewan-keith/rstansim")
library(parallel)
library(patchwork)
library(ggrepel)
library(pbmcapply)
library(pbapply)
library(RColorBrewer)
source("../R/load_functions.R")

year = 365.25
param_colors = brewer.pal(2, "Set2")
names(param_colors) = c("lambda", "phi")
n_chains = 4

# Initialise values for verification runs
data_null = list(
  t0 = -5*year,
  alpha = 1,
  beta = 1,
  relapse_clinical_immunity = 0,
  gamma_d = 1/434.,
  gamma_l = 1/223,
  delta = 1/162,
  phi = 0,
  f = 1/72,
  r = 1/60,
  p_long = 0,
  p_silent = 0,
  N = 10000,
  population_size = 10000,
  
  n_dormant = 4,
  eps = 1,
  kappa = 1,
  phase = 130
)

model_file = "../stan/temperate_9.stan"
model = stan_model(model_file)

my_state_init = state_init_3
my_calculate_quantities = calculate_quantities_3
```

We construct a scenario by hand that appears realistic for Chinese temperate data:

```{r}
dt = year/12
ts = seq(1, 5*year, by=dt)
n_times = length(ts)
cases = rpois(n_times, 100)
data = data_init(data_null,
                 n_times = n_times,
                 ts = ts,
                 cases = cases,
                 alpha = 0.6,
                 beta = 0.2,
                 p_long = 1,
                 p_silent = 0,
                 eps = 0,
                 kappa = 3)
data$y0 = my_state_init(data, I0=0.01)

true_params = list(lambda = 0.05, phi_inv = 0.1, relapse_clinical_immunity = 0.9, eps=0.01, kappa=3, phase=120)

sim_out = my_simulate_data_list(
  file = model_file,
  data_name = "dummy_data",
  input_data = data,
  param_values = true_params,
  vars = c("y")
)$y %>%
  as_tibble() %>%
  setNames(names(data$y0)) %>%
  mutate(time = data$ts, .before=0) %>%
  my_calculate_quantities(data)

plot_data = sim_out %>%
  pivot_longer(-time, names_to="Compartment", values_to="Proportion") %>%
  mutate(Type = ifelse(Compartment %>% str_detect("Clinical|All"), "Incidence", "Compartment")) %>%
  filter(Compartment %in% c("Infectious", "Latent", "Dormant"))
plot_labels = plot_data %>%
  filter(time/year > 48) %>%
  group_by(Compartment) %>%
  arrange(desc(Proportion)) %>%
  slice(1) %>%
  filter(Proportion > 1e-12)

ggplot(plot_data, aes(x=time/year, y=Proportion, color=Compartment)) +
  geom_line() +
  geom_label_repel(data=plot_labels, aes(label=Compartment), direction="y", nudge_x=0.01, max.overlaps=1e6, fill = alpha(c("white"),0.3), label.padding=0.1, alpha=0.7) +
  # ggforce::facet_zoom(xlim = c(max(data$ts)/year - 2)) +
  coord_cartesian(ylim=c(0, NA)) +
  labs(subtitle = "Manual parameters")

plot_data = sim_out %>%
  pivot_longer(-time, names_to="Compartment") %>%
  filter(Compartment %>% str_detect("Clinical|All")) %>%
  mutate(Type = Compartment %>% str_extract("Clinical|All"),
         Compartment = Compartment %>% str_remove("Clinical|All")) %>%
  drop_na()
plot_labels = plot_data %>%
  # filter(time/year > 48) %>%
  group_by(Compartment, Type) %>%
  arrange(desc(value)) %>%
  slice(1) %>%
  filter(value > 1e-12)

ggplot(plot_data, aes(x=time/year, y=value, color=Compartment)) +
  geom_line() +
  geom_label_repel(data=plot_labels, aes(label=Compartment), direction="y", nudge_x=0.01, max.overlaps=1e6, fill = alpha(c("white"),0.3), label.padding=0.1, alpha=0.7) +
  facet_grid(rows = vars(Type)) +
  # ggforce::facet_zoom(xlim = c(48, 50)) +
  coord_cartesian(ylim=c(0, NA)) +
  # theme(legend.position = "none") +
  labs(subtitle = "Manual parameters")
```

What happens when we simulate data?

```{r}
data_sim = data

synth_df = my_simulate_data(
    file = model_file,
    path = "sim_data",
    data_name = "nonstationary",
    input_data = data_sim,
    param_values = true_params,
    vars = c("ts", "incidence", "sim_cases")
  )


ggplot(synth_df, aes(x=ts/year, y=cases)) +
  geom_col()
```

How well does parameter recovery work?

```{r}
data_recovery = data_init(data_sim,
                          n_times = nrow(synth_df),
                          ts = synth_df$ts,
                          cases = synth_df$cases)

tictoc::tic()
n_chains = 2
samp = sampling(model,
                data = data_recovery,
                iter = 100,
                chains = n_chains,
                # init = rep(list(.theta_init), n_chains), # Start from MLE solution
                cores = n_chains,
                pars = "y_extended",
                include = FALSE)
tictoc::toc()
```

```{r}
# inspect parameter posteriors
parameters = c("lambda", "phi_inv", "eps", "kappa", "phase", "relapse_clinical_immunity")
posterior_seasonal = rstan::extract(samp, parameters[parameters %in% names(samp)]) %>%
  as_tibble() %>%
  mutate(phi = 1/phi_inv) %>%
  select(-phi_inv) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value")

true_params_df = as_tibble(true_params) %>%
  mutate(phi = 1/phi_inv) %>%
  select(-phi_inv) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value")

posterior_seasonal %>%
  ggplot(aes(x = value, fill=parameter, color=parameter)) +
  geom_vline(data=true_params_df, aes(xintercept = value), linetype="dashed") +
  geom_density(alpha=0.5) +
  scale_fill_manual(values = param_colors, drop=F) +
  scale_color_manual(values = param_colors, drop=F) +
  coord_cartesian(xlim = c(0, NA)) +
  facet_wrap(vars(parameter), scales="free", labeller=plot_labeller_novar) +
  labs(subtitle = "Parameter estimates")

# ggsave("plots/rondonia_seasonal_posterior.png", width=8, height=4)

# Plot credible intervals
n_traces = 100
incidence = rstan::extract(samp, "incidence")[[1]]
sample_ix = sample(seq_len(dim(incidence)[1]), n_traces, replace=T)
incidence_sample_nonseasonal = as_tibble(t(incidence[sample_ix,])) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
  drop_na(j) %>%
  group_by(j) %>%
  mutate(ts = data_recovery$ts[j],
         lower = quantile(incidence, 0.025, na.rm=T),
         upper = quantile(incidence, 0.975, na.rm=T),
         legend = "95% prediction interval")


ggplot(mapping = aes(x=ts/years)) +
  geom_ribbon(data = incidence_sample_nonseasonal,
              aes(ymin=lower, ymax=upper, fill="95% mean incidence interval"),
              alpha = 0.5) +
  geom_point(data = tibble(ts = data_recovery$ts, cases = data_recovery$cases),
             aes(y = cases, group = NULL, color="Incidence data")) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_fill_manual("", 
                    breaks = c("95% case prediction interval", "95% mean incidence interval", "Incidence data"),
                    values = c("steelblue", "tomato", "black")) +
  scale_colour_manual("", 
                      breaks = c("95% case prediction interval", "95% mean incidence interval", "Incidence data"),
                      values = c("steelblue", "tomato", "black")) +
  labs(subtitle = "Seasonal model, monthly synthetic data",
       x = "Year",
       y = "Annual incidence")
```

What if we refit the model not using correct parameters?

```{r}
# We reuse temperate model 6 and set it to have no clinical immunity
data_recovery_2 = data_init(data_recovery,
                            relapse_clinical_immunity = 0)
tictoc::tic("Fitting relapse clinical immunity")
samp_noimmunity = sampling(stan_model("../stan/temperate_6_1.stan"),
                           data = data_recovery_2,
                           iter = 100,
                           chains = n_chains,
                           # init = rep(list(theta_init), n_chains), # Start from MLE solution
                           cores = n_chains)
tictoc::toc()
```

```{r}
# inspect parameter posteriors
posterior_seasonal = rstan::extract(samp_noimmunity, c("lambda", "phi_inv")) %>%
  as_tibble()

posterior_seasonal %>%
  mutate(phi = 1/phi_inv,
         phi_inv = NULL) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  ggplot(aes(x = value, fill=parameter, color=parameter)) +
  geom_density(alpha=0.5) +
  scale_fill_manual(values = param_colors, drop=F) +
  scale_color_manual(values = param_colors, drop=F) +
  coord_cartesian(xlim = c(0, NA)) +
  facet_wrap(vars(parameter), scales="free") +
  labs(subtitle = "Parameter estimates")

# Plot credible intervals
n_traces = 100
incidence = rstan::extract(samp_noimmunity, "incidence")[[1]]
sample_ix = sample(seq_len(dim(incidence)[1]), n_traces, replace=T)
plot_data_noimmunity = as_tibble(t(incidence[sample_ix,])) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
  drop_na(j) %>%
  mutate(ts = data_recovery_2$ts[j])

ggplot(mapping = aes(x=ts)) +
  geom_line(data = plot_data_noimmunity,
            aes(y=incidence, color=paste(n_traces, "traces of mean incidence"), group=trace),
            alpha = 0.1) +
  geom_point(data = tibble(ts = data_recovery$ts, cases = data_recovery$cases),
             aes(y = cases, group = NULL, color="Case data")) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_fill_manual("", 
                    breaks = c("95% case prediction interval", paste(n_traces, "traces of mean incidence"), "Case data"),
                    values = c("steelblue", "tomato", "black")) +
  scale_colour_manual("", 
                      breaks = c("95% case prediction interval",paste(n_traces, "traces of mean incidence"), "Case data"),
                      values = c("steelblue", "tomato", "black")) +
  labs(title = "Seasonal model, monthly data",
       subtitle = "Refit where relapse_clinical_immunity = 0",
       x = "Year",
       y = "Annual incidence")
```