_---
title: "PLAN for Chapter 1. Modelology for fitting ODE epidemic models"
subtitle: "This RMarkdown is a template as initially outlined in the progress report. Each task will be outlined and then followed by a code block to be completed."
output:
html_notebook:
number_sections: true
---

**This should be similar to `Chapter_01.Rmd` but with the experiment data structure improved to take advantage of more cores than the number of scenarios.**

# Introduction

When calibrating their vivax population transmission models, White and Champagne lack longitudinal time series data and are forced to make equilibrium assumptions at a point in time (initially from Griffin from his 2014 P.f model). Here, derivatives are set to zero and the equations solved for unknowns. In the case of Champagne, the transmission rate $\lambda$ so that force of infection is $\lambda$ multiplied by the sum of infectious compartments $\sum{I}$.

This assumption does not hold for areas with moderate to strong seasonality and long-term trends. This document demonstrates the standard implementation of dynamic ODE parameter-fitting methods applied to the White/Champagne style model and extends the form of the model (and its corresponding parameter estimation routine) to provide increasingly flexible relaxations to the original uses of the White/Champagne model. The end result will be a model and fitting routine that is flexible enough to reflect vivax epidemics in non-stationary regions, unlike the original implementation.

[...]

# Models

Stan and posterior sampling will be used for all analyses.

```{r setup}
library(R.utils)
library(tidyverse)
library(rstan)
library(rstansim) # devtools::install_github("ewan-keith/rstansim")
library(parallel)
library(patchwork)
library(pbmcapply)
library(pbapply)
library(memoise)
library(RColorBrewer)
source("../R/constants.R")
source("../R/load_functions.R")

model_champagne2022_seasonal_ext = "../stan/champagne2022_seasonal_ext_2.stan"
stan_model_champagne2022_seasonal_ext = stan_model(model_champagne2022_seasonal_ext)
model_champagne2022 = "../stan/champagne2022.stan"
stan_model_champagne2022 = stan_model(model_champagne2022)

n_cores = parallelly::availableCores()
options(mc.cores = n_cores)
message("Running on ", n_cores, " cores")
rstan_options(auto_write = TRUE)

n_iter = 500 # should be at least 500
n_chains = 8
n_repetitions = 1 # how many times to duplicate each scenario
cores_per_sampler = n_chains # 1 for mclapply, or n_chains if not running lots of scenarios
n_traces = 200 # limit for plotting sample trajectories
```

## Model implementation

We begin with extending Champagne's 2022 model for tropical vivax to include seasonality.

First we implement and verify the parameter recovery ability of Stan with the Champagne model as published in 2022.

```{r}
#initial conditions
I_init = 0.01
y0 = c(Il=0, I0=I_init, Sl=0, S0=1-I_init, CumulativeInfections=0)

# constants for Stan
data_consts = list(y0 = y0,
                   t0 = -50*years,
                   N = 1.8e6/100, # population size,
                   r = 1./60, # r
                   gammal = 1./223, # gammal
                   f = 1./72, # f
                   alpha = 0.8, # alpha,
                   beta = 0.429/0.8, # beta, Nekkab 2021
                   delta = 0
)
```

## Parameter recovery on real data

We test whether parameter recovery works on real datasets. We will find data from a variety of settings (e.g., transmission levels, remoteness, strains) to demonstrate generalisability.

Data:

- Hainan data (tropical China) but this has been difficult to acquire.
We will not be investigating Chinese Yunnan (southern mountainous) or Henan (central temperate) data which we do have because the Yunnan strain is not known to be tropical, and the Henan data is temperate (our current temperate model does not align with this data convincingly).
- Brazilian 'integrated data set', available per county or municipality and very detailed.

```{r}
library(MalariaData)

integrated_dataset %>%
  count(notification.date) %>%
  ggplot(aes(x = notification.date, y = n)) +
  geom_col() +
  labs(title = "Lab-confirmed P. vivax cases in Brazil",
       x = "Month",
       y = "Count")

ggsave("../plots/brazil_cases.png", width=8, height=2)
```

Define our usable data

```{r}
min_year = 2015
max_year = 2018

rondonia.all = integrated_dataset %>%
  filter(exam.result == "vivax",
         infection.state == "RO") %>%
  count(notification.year, notification.date, name = "cases") %>%
  mutate(include = case_when(notification.year >= min_year & notification.year <= max_year ~ "grey20",
                             notification.year <= max_year ~ "grey40",
                             TRUE ~ "grey")) %>%
  # mutate(include = ifelse(notification.year >= min_year & notification.year <= max_year, "grey20", "grey")) %>%
  as_tibble()

yearly_counts = rondonia.all %>%
  group_by(notification.year) %>%
  summarise(cases = sum(cases) / annual_subdivisions) %>%
  mutate(notification.date = as.Date(paste0(notification.year, "-01-01"))) %>%
  bind_rows(tail(., 1) %>% mutate(notification.date = notification.date + years(1)))

ggplot(rondonia.all, aes(x = notification.date, y=cases, fill=include)) +
  annotate("rect", xmin=as.Date(paste0(min_year, "-01-01")), xmax=as.Date(paste0(max_year, "-12-31")), ymin=-Inf, ymax=Inf,
           fill = "blue", alpha = 0.15) +
  geom_col() +
  geom_step(data=yearly_counts, aes(fill=NULL), color="tomato") +
  coord_cartesian(ylim = c(0, NA)) +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  scale_y_continuous(name = "Monthly cases",
                     labels = label_auto2,
                     sec.axis = sec_axis(trans=~.*annual_subdivisions, name="Annual cases", labels=label_auto2)) +
  scale_fill_identity() +
  labs(title = "Selected P. vivax cases infected in Rondonia",
       x = "Notification date",
       y = "Cases") +
  theme(axis.ticks.y.right = element_line(color = "tomato"),
        axis.text.y.right = element_text(color = "tomato"),
        axis.title.y.right = element_text(color = "tomato"))

ggsave("../plots/rondonia_selected.png", width=8, height=2)

rondonia = rondonia.all %>%
  filter(include == "grey20")
```

```{r}
source("../R/functions/decompose_data.R")
rondonia_decomp = rondonia %>%
  filter(notification.year >= min_year, notification.year <= max_year) %>%
  dplyr::rename(Date = notification.date,
                Cases = cases) %>%
  mutate(logCases = log(Cases)) %>%
  decompose_data()

rondonia_decomp %>%
  filter(name %in% c("Cases", "trend", "seasonal", "remainder")) %>%
  ggplot(aes(x = Date, y = value, color = value)) +
  geom_line() +
  coord_cartesian(ylim = c(0, NA)) +
  facet_wrap(vars(name), scales="free_y")
```

Create data for the full model

```{r}
data_ro = data_init(data_consts,
                    cases = rondonia$cases,
                    ts = seq_len(nrow(rondonia)) * years / annual_subdivisions,
                    n_times = nrow(rondonia))
```

Create data for the nonseasonal model

```{r}
rondonia_agg = integrated_dataset %>%
  filter(exam.result == "vivax",
         infection.state == "RO",
         notification.year >= min_year,
         notification.year <= max_year) %>%
  mutate(notification.date = as.Date(paste(notification.year, "01", "01", sep="-")) + years(1) - days(1)) %>%
  count(notification.date, name = "cases")

# Aggregate data to a yearly level
data_ro_agg = data_init(data_consts,
                        cases = rondonia_agg$cases,
                        ts = seq_len(nrow(rondonia_agg)) * years,
                        n_times = nrow(rondonia_agg))

tibble(ts = data_ro_agg$ts / years + (min_year-1),
       cases = data_ro_agg$cases) %>%
  ggplot(aes(x = ts, y = cases)) +
  geom_line() +
  coord_cartesian(ylim = c(0, NA))
```

Perform MCMC sampling for both models

```{r}
tictoc::tic("Seasonal model elapsed")
samp_ext = sampling(stan_model_champagne2022_seasonal_ext,
                    data = data_ro,
                    iter = n_iter,
                    chains = n_chains)
tictoc::toc()

tictoc::tic("Nonseasonal model elapsed")
samp_ext_nonseasonal = sampling(stan_model_champagne2022,
                                data = data_ro_agg,
                                iter = n_iter,
                                chains = n_chains)
tictoc::toc()
```


```{r}
# inspect parameter posteriors
posterior_seasonal = rstan::extract(samp_ext, c("lambda", "phi_inv", "eps", "kappa", "phase")) %>%
  as_tibble() %>%
  mutate(phi = 1/phi_inv) %>%
  select(-phi_inv)

posterior_seasonal_summary = posterior_seasonal %>%
  mutate(peak_day = years/4 + phase) %>%
  pivot_longer(everything(), names_to = "parameter") %>%
  group_by(parameter) %>%
  summarise(mean = mean(value),
            lq = quantile(value, 0.025),
            uq = quantile(value, 0.975))

# Plot credible intervals
incidence = rstan::extract(samp_ext, "incidence")[[1]]
incidence_sample_nonseasonal = as_tibble(t(incidence)) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
  group_by(j) %>%
  mutate(ts = data_ro$ts[j],
         lower = quantile(incidence, 0.025, na.rm=T),
         upper = quantile(incidence, 0.975, na.rm=T),
         legend = "95% prediction interval")

sim_cases = rstan::extract(samp_ext, "sim_cases")[[1]]
sim_cases_sample_nonseasonal = as_tibble(t(sim_cases)) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "sim_cases") %>%
  group_by(j) %>%
  mutate(ts = data_ro$ts[j],
         lower = quantile(sim_cases, 0.025, na.rm=T),
         upper = quantile(sim_cases, 0.975, na.rm=T),
         legend = "95% prediction interval") %>%
  distinct(j, .keep_all=T)

p_seasonal_ci = ggplot(mapping = aes(x=ts + as.Date(paste0(min_year, "-01-01"))-days(1))) +
  geom_ribbon(data = sim_cases_sample_nonseasonal,
              aes(ymin=lower, ymax=upper, fill="95% case prediction interval"),
              alpha = 0.5) +
  geom_ribbon(data = incidence_sample_nonseasonal,
              aes(ymin=lower, ymax=upper, fill="95% mean incidence interval"),
              alpha = 0.5) +
  geom_point(data = tibble(ts = data_ro$ts, cases = data_ro$cases),
             aes(y = cases, group = NULL, color="Case data")) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_fill_manual("", 
                    breaks = c("95% case prediction interval", "95% mean incidence interval", "Case data"),
                    values = c("steelblue", "tomato", "black")) +
  scale_colour_manual("", 
                      breaks = c("95% case prediction interval", "95% mean incidence interval", "Case data"),
                      values = c("steelblue", "tomato", "black")) +
  labs(title = "Posterior incidence estimates for Rondonia",
       subtitle = "Estimated using the seasonal model at periodic equilibrium",
       x = "Year",
       y = "Monthly incidence")
p_seasonal_ci

ggsave("../plots/rondonia_seasonal_incidence.png", width=8, height=4)
```

And, non-seasonal

```{r}
# inspect parameter posteriors
posterior_nonseasonal = rstan::extract(samp_ext_nonseasonal, c("lambda", "phi_inv")) %>%
  as_tibble() %>%
  mutate(phi = 1/phi_inv) %>%
  select(-phi_inv)

# Plot traces
incidence = rstan::extract(samp_ext_nonseasonal, "incidence")[[1]]
sample_ix = sample(seq_len(dim(incidence)[1]), n_traces, replace=T)
incidence_sample_nonseasonal = as_tibble(t(incidence[sample_ix,])) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
  mutate(ts = data_ro_agg$ts[j],
         lower = quantile(incidence, 0.025, na.rm=T),
         upper = quantile(incidence, 0.975, na.rm=T),
         legend = "95% prediction interval")

sim_cases = rstan::extract(samp_ext_nonseasonal, "sim_cases")[[1]]
sim_cases_sample_nonseasonal = as_tibble(t(sim_cases[sample_ix,])) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "sim_cases") %>%
  mutate(ts = data_ro_agg$ts[j],
         lower = quantile(sim_cases, 0.025, na.rm=T),
         upper = quantile(sim_cases, 0.975, na.rm=T),
         legend = "95% prediction interval") %>%
  distinct(ts, .keep_all=T)

p_ns_ci = ggplot(mapping = aes(x=ts + as.Date(paste0(min_year, "-01-01"))-years(1))) +
  geom_ribbon(data = sim_cases_sample_nonseasonal,
              aes(ymin=lower, ymax=upper, fill="95% case prediction interval"),
              alpha = 0.5) +
  geom_ribbon(data = incidence_sample_nonseasonal,
              aes(ymin=lower, ymax=upper, fill="95% mean incidence interval"),
              alpha = 0.5) +
  geom_point(data = tibble(ts = data_ro_agg$ts, cases = data_ro_agg$cases),
             aes(y = cases, group = NULL, color="Incidence data")) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_fill_manual("", 
                    breaks = c("95% case prediction interval", "95% mean incidence interval", "Incidence data"),
                    values = c("steelblue", "tomato", "black")) +
  scale_colour_manual("", 
                      breaks = c("95% case prediction interval", "95% mean incidence interval", "Incidence data"),
                      values = c("steelblue", "tomato", "black")) +
  labs(subtitle = "Non-seasonal model, annual data",
       x = "Year",
       y = "Annual incidence")
p_ns_ci

# ggsave("plots/rondonia_ns_ci.png", width=8, height=4)
```

Generate subplots

Comparison between posteriors

```{r}
posterior_comparison = bind_rows(Nonseasonal = posterior_nonseasonal,
                                 Seasonal = posterior_seasonal,
                                 .id = "Model") %>%
  pivot_longer(-Model, names_to="parameter") %>%
  mutate(parameter = fct_relevel(parameter, c("lambda", "phi", "eps", "kappa", "phase")))

posterior_ranges_2 = posterior_comparison %>%
  group_by(parameter) %>%
  summarise(mean = mean(value, na.rm=T),
            lq = quantile(value, 0.01, na.rm=T),
            uq = quantile(value, 0.99, na.rm=T),
            min = lq - (mean-lq)*1.5,
            max = uq + (uq-mean)*1.5,
            Model = NA) %>%
  mutate(min = case_when(parameter %in% c("phi", "phase", "eps", "kappa") ~ 0,
                         TRUE ~ min),
         max = case_when(parameter == "phase" ~ years,
                         parameter == "eps" ~ 1,
                         TRUE ~ max)) %>%
  pivot_longer(c(min, max))

ggplot(posterior_comparison, aes(x = value, fill = Model, color=Model)) +
  geom_blank(data = posterior_ranges_2) +
  geom_density(alpha = 0.5) +
  facet_wrap(vars(parameter), scales="free", labeller=plot_labeller_novar) +
  labs(title = "Posterior comparison between seasonal and nonseasonal models",
       x = "Value",
       y = "Posterior density") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

ggsave("../plots/rondonia_comparison.png", width=8, height=4)
```

```{r}
(p_seasonal_ci + labs(title = NULL, subtitle = "Seasonal model")) +
  (p_ns_ci + labs(title = NULL, subtitle = "Nonseasonal model")) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Posterior model fits",
                  tag_levels = "A")
ggsave("../plots/rondonia_incidence.png", width=8, height=4)
```

Examine the estimated seasonal pattern

```{r}
omega = function(t, eps, kappa, phase=365.25/2) {
  return(eps + (1-eps)*pi/beta(0.5, kappa+0.5)*((1+sin(2*pi*(t - phase)/365.25))/2)^kappa)
}

omega_params = rstan::extract(samp_ext, c("eps", "kappa", "phase")) %>%
  as_tibble() %>%
  mutate(i = row_number())

tt = seq(0, years, length.out=101)
omega_ts = lapply(seq_len(nrow(omega_params)), function(i) {
  omega(tt, omega_params$eps[i], omega_params$kappa[i], omega_params$phase[i])
})
omega_params$omega = omega_ts
omega_params$t = rep(list(tt), nrow(omega_params))
omega_params %>%
  unnest(c(t, omega)) %>%
  ggplot(aes(x = t/years*annual_subdivisions, y=omega, group=i)) +
  geom_line(alpha=0.05, color="grey10") +
  geom_hline(yintercept = 1, linetype="dashed") +
  coord_cartesian(ylim = c(0, NA)) +
  scale_x_continuous(breaks = 0:annual_subdivisions, expand = c(0,0),
                     labels = c(month.abb, month.abb[1])) +
  labs(title = paste("Posterior estimate of ω(t) for Rondonia P. vivax between", min_year, "and", max_year),
       x = "Month",
       y = "ω(t)")

ggsave("../plots/rondonia_omega.png", width=8, height=4)
```

Comparison of stationary and nonstationary parameter estimates

```{r}
posterior_comparison = bind_rows(Seasonal = posterior_seasonal,
                                 Nonseasonal = posterior_nonseasonal,
                                 .id = "Model") %>%
  pivot_longer(-Model, names_to="parameter") %>%
  filter(parameter %in% c("lambda", "phi"))

posterior_ranges_2 = posterior_comparison %>%
  group_by(parameter) %>%
  summarise(mean = mean(value),
            lq = quantile(value, 0.01),
            uq = quantile(value, 0.99),
            min = lq - (mean-lq)*1.5,
            max = uq + (uq-mean)*1.5,
            Model = NA) %>%
  pivot_longer(c(min, max))

ggplot(posterior_comparison, aes(x = value, fill = Model, color=Model)) +
  geom_blank(data = posterior_ranges_2) +
  geom_density(alpha = 0.5) +
  facet_wrap(vars(parameter), scales="free", labeller=plot_labeller_novar) +
  labs(title = "Posterior comparison between seasonal and nonseasonal models",
       subtitle = "Method applied to data from Rondonia between 2015 and 2018",
       x = "Value",
       y = "Posterior density")

ggsave("../plots/rondonia_seasonality_comparison.png", width=8, height=4)
```

What if we used the steady-state fit to simulate posteriors in the full model?

```{r}
n_samples = 500
composite_list = pbmclapply(seq_len(n_samples), function(i) {
  theta_composite = list(
    lambda = sample(posterior_nonseasonal$lambda, 1),
    phi_inv = 0.1, # This doesn't matter if not sampling cases sample(posterior_seasonal$phi_inv, 1)
    eps = sample(posterior_seasonal$eps, 1),
    kappa = sample(posterior_seasonal$kappa, 1),
    phase = sample(posterior_seasonal$phase, 1)
  )
  
  my_simulate_data(
    file = model_champagne2022_seasonal_ext,
    path = "sim_data",
    data_name = "rondonia_fit",
    input_data = data_ro,
    param_values = theta_composite,
    vars = c("ts", "incidence", "sim_cases")
  )
})
composite_list[sapply(composite_list, function(x) {"try-error" %in% class(x)})] <- NULL

incidence = rstan::extract(samp_ext, "incidence")[[1]]
ix = sample(seq_len(dim(incidence)[1]), n_traces, replace=T)
ts_sample = as_tibble(t(incidence[ix,])) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
  mutate(ts = data_ro$ts[j])

posterior_comparison = composite_list %>%
  bind_rows(.id = "trace") %>%
  mutate(method = "Nonseasonal calibration\nwith seasonality added") %>%
  bind_rows(ts_sample %>% mutate(method = "Seasonal model calibration", j=NULL)) %>%
  as_tibble()

ylimits = c(0, max(data_ro$cases) * 1.2)

posterior_comparison %>%
  ggplot(aes(x = ts + as.Date(paste0(min_year, "-01-01")), y = incidence)) +
  geom_line(aes(color=method, group=trace), alpha=0.1) +
  geom_point(data = tibble(ts = data_ro$ts, cases = data_ro$cases, color="Incidence data"), aes(y=cases)) +
  coord_cartesian(ylim = ylimits) +
  scale_colour_manual(NULL, 
                      breaks = c("Incidence data", "Seasonal model calibration", "Nonseasonal calibration\nwith seasonality added"),
                      values = c("black", "tomato", "steelblue")) +
  guides(color = guide_legend(override.aes = list(alpha = 1))) +
  labs(title = "Calibrated using the full seasonal model or the steady-state approximation",
       color = "Calibration routine",
       x = "Year",
       y = "Cases / mean incidence")

ggsave("../plots/rondonia_composite.png", width=8, height=4)
```

How does this fitted seasonality compare against Martens (Malaria Atlas Project)

```{r}
library(sf)
rondonia_file = "~/Documents/MalariaData/data-raw/brazil/rondonia/data_EPSG_4326/11MIE250GC_SIR.shp" # https://purl.stanford.edu/df159kw0955

rondonia_geom = st_read(rondonia_file) %>%
  st_union() %>%
  st_as_sf() %>%
  mutate(Name = "Rondonia")
```


```{r}
map_folder = "~/Documents/MalariaData/data-raw/TSI-Martens2-Pv.2000-2022.Data.5km.Data"

# map = read_map(map_folder)
# map_sites = extract_layers(map, sites, "month", FALSE)
tifs = tibble(path = list.files(map_folder, full.names=T)) %>%
  mutate(filename = basename(path),
         year = str_extract(filename, "(1|2)[0-9]{3}") %>% as.numeric()) %>%
  filter(year >= min_year, year <= max_year)

x = terra::rast(tifs$path)
```

```{r}
rondonia_map = terra::extract(x, rondonia_geom) %>%
  pivot_longer(-ID, names_to="layer") %>%
  group_by(ID, layer) %>%
  summarise(value = mean(value, na.rm=T),
            .groups = "drop") %>%
  mutate(Name = rondonia_geom$Name[ID],
         year = layer %>% str_extract("[0-9]{4}") %>% as.numeric(),
         month = layer %>% str_extract("\\.[0-9]{2}\\.") %>% str_extract("[0-9]+") %>% as.numeric(),
         yearmonth = year*12 + month,
         date = as.Date(paste(year, month, "15", sep="-")))

rondonia_map_mean = rondonia_map %>%
  group_by(month) %>%
  summarise(value = mean(value))

ggplot(rondonia_map, aes(x=month, y=value, color=year, group=year)) +
  geom_line() +
  geom_line(data=rondonia_map_mean, aes(color=NULL, group=NULL), linetype="dashed") +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(limits = 0:1)
```

Can we perform the fit using the nonstationary model?

Assemble data

```{r}
rondonia_ns = rondonia.all %>%
  filter(notification.year <= max_year)

data_ro_ns = data_init(data_consts,
                       cases = rondonia_ns$cases,
                       ts = seq_len(nrow(rondonia_ns)) * years / annual_subdivisions,
                       n_times = nrow(rondonia_ns))
```

Perform MCMC sampling for the nonstationary model

```{r}
tictoc::tic("Seasonal model elapsed")
samp_ns = sampling(stan_model("../stan/champagne2022_nonstationary_2.stan"),
                   data = data_ro_ns,
                   iter = n_iter,
                   chains = n_chains)
tictoc::toc()
```

Inspect the fit:

```{r}
posterior_nonstationary = rstan::extract(samp_ns, c("lambda", "phi_inv", "eps", "kappa", "phase", "tstar", "xi")) %>%
  as_tibble() %>%
  mutate(phi = 1 / phi_inv,
         tstar = tstar / years + 2009) %>%
  select(-phi_inv)

posterior_nonstationary_summary = posterior_nonstationary %>%
  mutate(peak_day = years/4 + phase) %>%
  pivot_longer(everything(), names_to = "parameter") %>%
  group_by(parameter) %>%
  summarise(mean = mean(value),
            lq = quantile(value, 0.025),
            uq = quantile(value, 0.975))

alpha = 0.05
lower_name = paste0(as.character(100*alpha/2), "%")
upper_name = paste0(as.character((100*(1-alpha/2))), "%")
posterior_credints = posterior_nonstationary %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarise(mean = mean(value),
            lower = quantile(value, alpha/2),
            upper = quantile(value, 1-alpha/2)) %>%
  mutate_if(is.numeric, function(x) {as.character(signif(x, 3))}) %>%
  dplyr::rename(Variable = name,
                !!lower_name := lower,
                !!upper_name := upper)

posterior_ranges = tribble(
  ~parameter, ~min, ~max,
  "lambda", 0, max(posterior_nonstationary$lambda * 1.5),
  "phi", 0, max(posterior_nonstationary$phi * 1.5),
  "eps", 0, 1,
  "kappa", 0, max(posterior_nonstationary$kappa * 1.5),
  "phase", 0, years,
  "tstar", 2009, 2019,
  "xi", 0, 1
) %>%
  pivot_longer(c(min, max))

# Plot posterior densities
posterior_nonstationary %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  ggplot(aes(x = value, fill=parameter, color=parameter)) +
  geom_blank(data = posterior_ranges, ) +
  geom_density(alpha=0.5) +
  scale_fill_manual(values = param_colors, drop=F) +
  scale_color_manual(values = param_colors, drop=F) +
  # coord_cartesian(xlim = c(0, NA)) +
  facet_wrap(vars(parameter), scales="free", labeller = plot_labeller_novar) +
  labs(title = "Posterior parameter estimates for Rondonia",
       subtitle = "Estimated using the nonstationary seasonal model",
       x = "Value",
       y = "Posterior density") +
  theme(legend.position = "none")

ggsave("../plots/rondonia_nonstationary_posterior.png", width=8, height=4)


sim_cases = rstan::extract(samp_ns, "sim_cases")[[1]]
sim_cases_sample_nonstationary = as_tibble(t(sim_cases)) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "sim_cases") %>%
  group_by(j) %>%
  mutate(ts = data_ro_ns$ts[j],
         lower = quantile(sim_cases, 0.025, na.rm=T),
         upper = quantile(sim_cases, 0.975, na.rm=T),
         legend = "95% prediction interval") %>%
  distinct(j, .keep_all=T)

incidence = rstan::extract(samp_ns, "incidence")[[1]]
incidence_sample_nonstationary = as_tibble(t(incidence)) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
  drop_na(j) %>%
  group_by(j) %>%
  mutate(ts = data_ro_ns$ts[j],
         lower = quantile(incidence, 0.025, na.rm=T),
         upper = quantile(incidence, 0.975, na.rm=T),
         legend = "95% prediction interval")


ggplot(mapping = aes(x=ts / years + 2009)) +
  geom_ribbon(data = sim_cases_sample_nonstationary,
              aes(ymin=lower, ymax=upper, fill="95% case prediction interval"),
              alpha = 0.5) +
  geom_ribbon(data = incidence_sample_nonstationary,
              aes(ymin=lower, ymax=upper, fill="95% mean incidence interval"),
              alpha = 0.5) +
  geom_point(data = tibble(ts = data_ro_ns$ts, cases = data_ro_ns$cases),
             aes(y = cases, group = NULL, color="Case data")) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_x_continuous(breaks = 2009:2019) +
  scale_fill_manual("", 
                    breaks = c("95% case prediction interval", "95% mean incidence interval", "Case data"),
                    values = c("steelblue", "tomato", "black")) +
  scale_colour_manual("", 
                      breaks = c("95% case prediction interval", "95% mean incidence interval", "Case data"),
                      values = c("steelblue", "tomato", "black")) +
  labs(subtitle = "Seasonal model, monthly synthetic data",
       x = "Day",
       y = "Monthly incidence") +
  labs(title = "Posterior incidence estimates for Rondonia",
       subtitle = "Estimated using the nonstationary seasonal model",
       x = "Year",
       y = "Monthly incidence")

ggsave("../plots/rondonia_nonstationary_incidence.png", width=8, height=4)
```

Can we compare the posteriors between stationary/nonstationary?

```{r}
posterior_comparison = bind_rows(Stationary = posterior_seasonal,
                                 Nonstationary = posterior_nonstationary %>%
                                   mutate(lambda = lambda * xi),
                                 .id = "Model") %>%
  pivot_longer(-Model, names_to="parameter") %>%
  mutate(parameter = ifelse(parameter == "lambda", "λ(t>t*)", parameter)) %>%
  mutate(parameter = fct_relevel(parameter, c("λ(t>t*)", "phi", "xi", "tstar", "eps", "kappa", "phase")))

posterior_ranges_2 = posterior_comparison %>%
  group_by(parameter) %>%
  summarise(mean = mean(value, na.rm=T),
            lq = quantile(value, 0.01, na.rm=T),
            uq = quantile(value, 0.99, na.rm=T),
            min = lq - (mean-lq)*1.5,
            max = uq + (uq-mean)*1.5,
            Model = NA) %>%
  mutate(min = case_when(parameter %in% c("phi", "phase", "eps", "kappa") ~ 0,
                         TRUE ~ min),
         max = case_when(parameter == "phase" ~ years,
                         parameter == "eps" ~ 1,
                         TRUE ~ max)) %>%
  pivot_longer(c(min, max))

ggplot(posterior_comparison, aes(x = value, fill = Model, color=Model)) +
  geom_blank(data = posterior_ranges_2) +
  geom_density(alpha = 0.5) +
  facet_wrap(vars(parameter), scales="free", labeller=plot_labeller_novar) +
  labs(title = "Posterior comparison between stationary and nonstationary models",
       x = "Value",
       y = "Posterior density") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

ggsave("../plots/rondonia_nonstationary_comparison.png", width=8, height=4)
```

End of script.

```{r}
print("Done.")
```
