_---
title: "PLAN for Chapter 2. Simulation-estimation study for the temperate model"
subtitle: "This RMarkdown is a template as initially outlined in the progress report. Each task will be outlined and then followed by a code block to be completed."
output:
html_notebook:
number_sections: true
---



```{r setup}
library(R.utils)
library(tidyverse)
library(rstan)
library(rstansim) # devtools::install_github("ewan-keith/rstansim")
library(parallel)
library(patchwork)
library(pbmcapply)
library(pbapply)
library(memoise)
library(RColorBrewer)
source("../R/constants.R")
source("../R/load_functions.R")

n_cores = parallelly::availableCores()
# n_cores = 9
options(mc.cores = n_cores)
message("Running on ", n_cores, " cores")
rstan_options(auto_write = TRUE)

n_years = 5
n_iter = 200 # should be at least 500
n_chains = 4
n_repetitions = 1 # how many times to duplicate each scenario
limit_runs = Inf # set to a finite number for testing, or Inf to run all
timelimit_per_run = 60*60 * 5
n_traces = 200 # limit for plotting sample trajectories

data_null = list(
  t0 = -10*years,
  alpha = 1,
  beta = 1,
  relapse_clinical_immunity = 0,
  gamma_d = 1/434.,
  gamma_l = 1/223,
  delta = 1/162,
  phi = 0,
  f = 1/72,
  r = 1/60,
  p_long = 0,
  p_silent = 0,
  N = 10000,
  population_size = 10000,
  n_dormant = 4,
  eps = 1,
  kappa = 1,
  phase = 130
)
data_null$y0 = state_init_3(data_null, I0=0.01)
data_consts = data_null

# Compile model before we reach any sampling
model_file = "../stan/temperate_8_1.stan"
model = stan_model("../stan/temperate_8_1.stan")
```

## Model implementation

# Can we make a variety of scenarios?
Define a function to simulate cases for a scenario

```{r}
simulate_cases = function(data_consts, lambda, phi_inv, relapse_clinical_immunity) {
  n_times = annual_subdivisions * n_years
  data = data_init(data_consts,
                   n_times = n_times,
                   ts = seq_len(n_times) * years / annual_subdivisions,
                   cases = rep(1, n_times), # Dummy value
                   alpha = 0.6,
                   beta = 0.2,
                   p_long = 1,
                   p_silent = 0,
                   eps = 0,
                   kappa = 3)
  params = list(lambda=lambda, phi_inv=phi_inv, relapse_clinical_immunity=relapse_clinical_immunity)
  synth_df = my_simulate_data(
    file = model_file,
    path = "sim_data",
    data_name = "temperate_8_1",
    input_data = data,
    param_values = params,
    vars = c("ts", "incidence", "sim_cases")
  )
}
```

Create scenarios

```{r}
# Define the scenarios
scenarios = expand_grid(
  lambda = 0.015,
  phi_inv = 0.1,
  relapse_clinical_immunity = c(0.1, 0.9),
  rep = seq_len(n_repetitions)
) %>%
  mutate(ID = row_number(), .before=0)

# Simulate cases for each scenario
cases_scenarios = lapply(seq_len(nrow(scenarios)), function(i) {
  lambda = scenarios$lambda[i]
  phi_inv = scenarios$phi_inv[i]
  relapse_clinical_immunity = scenarios$relapse_clinical_immunity[i]
  simulate_cases(data_null, lambda, phi_inv, relapse_clinical_immunity)
})

# Attach simulation results to the scenarios dataframe
scenarios$ts = lapply(cases_scenarios, function(x) {x$ts})
scenarios$cases = lapply(cases_scenarios, function(x) {x$cases})

if (n_repetitions > 1) {
  title = paste(n_repetitions, "repetitions of synthetic scenario data")
} else {
  title = "Synthetic scenario data"
}
scenarios %>%
  unnest(cols = c("ts", "cases")) %>%
  ggplot(aes(x = ts/years, y = cases, group=interaction(lambda, relapse_clinical_immunity, rep))) +
  geom_line(alpha = 0.6) +
  scale_x_continuous(breaks = 0:n_years) +
  # scale_y_log10(labels = label_auto2) +
  scale_color_gradient(trans = "log", breaks=10^seq(-5, 5)) +
  facet_grid(rows = vars(lambda),
             cols = vars(relapse_clinical_immunity),
             labeller = plot_labeller_years,
             scales = "free_y") +
  labs(title = title,
       x = "Year",
       y = "Monthly cases")

ggsave("../plots/temperate_data.png", width=8, height=4)
```

Define a function to perform the model fit, then apply it to each scenario

```{r}
# Define function that takes simulated data
my_sampling = function(data_consts, ts, cases, true_lambda, true_phi_inv, true_relapse_clinical_immunity) {
  n_times = length(ts)
  data = data_init(data_consts,
                   cases = cases,
                   ts = ts,
                   n_times = n_times)
  true_params = list(lambda = true_lambda,
                     phi_inv = true_phi_inv,
                     relapse_clinical_immunity = true_relapse_clinical_immunity)
  fit = sampling(model,
                 data = data,
                 iter = n_iter,
                 init = rep(list(true_params), 1),
                 chains = 1,
                 seed = 0)
}

# Apply function to simulated data
tictoc::tic("Running Stan sampling")
fits_scenarios = lapply(seq_len(nrow(scenarios)), function(i) {
  my_sampling(data_consts, scenarios$ts[[i]], scenarios$cases[[i]], scenarios$lambda[i], scenarios$phi_inv[i], scenarios$relapse_clinical_immunity[i])
})
tictoc::toc()
```

```{r}
# Extract parameters into dataframe
estimates_scenarios = lapply(fits_scenarios, function(x) {
  posterior = rstan::extract(x, c("lambda", "phi_inv", "relapse_clinical_immunity")) %>%
    as_tibble()
})

# Attach re-estimated parameters
scenarios$est_lambda = lapply(estimates_scenarios, function(x) {x$lambda})
scenarios$est_phi = lapply(estimates_scenarios, function(x) {1/x$phi_inv})
scenarios$phi = 1 / scenarios$phi_inv
scenarios$est_relapse_clinical_immunity = lapply(estimates_scenarios, function(x) {x$relapse_clinical_immunity})

# Attach posterior predictions
scenarios$est_incidence = lapply(fits_scenarios, function(x) {
  incidence = rstan::extract(x, "incidence")[[1]]
  if (n_traces < dim(incidence)[1]) {
    sample_ix = seq_len(dim(incidence)[1])
  } else {
    sample_ix = sample(seq_len(dim(incidence)[1]), n_traces, replace=T)
  }
  
  incidence_sample = as_tibble(t(incidence[sample_ix,])) %>%
    mutate(j = row_number()) %>%
    pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
    drop_na(j) %>%
    group_by(j) %>%
    mutate(ts = scenarios$ts[[1]][j],
           lower = quantile(incidence, 0.025, na.rm=T),
           upper = quantile(incidence, 0.975, na.rm=T),
           legend = "95% prediction interval")
})
```

Do diagnostic plots

```{r}
# Interesting parameter estimates
for (param in c("lambda", "phi", "relapse_clinical_immunity")) {
  plot_data = scenarios %>%
    unnest(!!paste0("est_", param)) %>% 
    mutate(value = get(paste0("est_", param)))
  true_values = plot_data %>%
    mutate(true_value = get(param)) %>%
    distinct(lambda, relapse_clinical_immunity, true_value)
  if (param == "lambda") {
    x_limits = plot_data$lambda[1] * c(0.9, 1.1)
    plot_data = plot_data %>%
      filter(value >= plot_data$lambda[1]*0.8,
             value <= plot_data$lambda[1]*1.2)
  } else {
    x_limits = c(NA, NA)
  }
  param_rename = make_greek(param)
  if (n_repetitions > 1) {
    title = paste("Posterior estimates of", param_rename, "under", n_repetitions, "repetitions")
  } else {
    title = paste("Posterior estimate of", param_rename)
  }
  p = ggplot(plot_data, aes(x = value, fill = factor(rep), color = factor(rep), group = rep)) +
    geom_vline(data=true_values, aes(xintercept = true_value), linetype="dashed") +
    geom_density(alpha = 0.5) +
    # facet_grid(rows = vars(lambda), cols = vars(relapse_clinical_immunity), labeller = plot_labeller, scales="free_y") +
    facet_wrap(vars(lambda, relapse_clinical_immunity), labeller = plot_labeller, scales="free_y") +
    scale_color_brewer(palette = "Set2") +
    scale_fill_brewer(palette = "Set2") +
    coord_cartesian(xlim = x_limits) +
    labs(title = title) +
    theme(legend.position = "none")
  print(p)
  ggsave(paste0("../plots/temperate_", param, ".png"), width=8, height=4)
}

```

Plot incidence fits

```{r}
incidence_scenarios = lapply(fits_scenarios, extract_incidence, scenarios$ts[[1]])
incidence_join = lapply(seq_along(incidence_scenarios), function(i) {
  incidence_scenarios[[i]] %>%
    mutate(ID = i)
}) %>%
  bind_rows() %>%
  left_join(scenarios %>% select(ID, lambda, relapse_clinical_immunity, rep), by="ID")

scenarios_overlay = scenarios %>%
  unnest(c(ts, cases))

max_y = max(scenarios_overlay$cases) * 1.2

ggplot(incidence_join, aes(x = ts, y = incidence, color=factor(rep), group = rep)) +
  geom_line(alpha = 0.25) +
  # geom_point(data=scenarios_overlay, aes(y=cases)) +
  # geom_vline(data=true_values, aes(xintercept = true_value), linetype="dashed") +
  # geom_density(alpha = 0.5) +
  facet_grid(rows = vars(lambda), cols = vars(relapse_clinical_immunity), labeller = plot_labeller_years) +
  # scale_fill_manual(values = param_colors, drop=F) +
  # scale_color_manual(values = param_colors, drop=F) +
  coord_cartesian(ylim = c(0, max_y)) +
  labs(title = "Estimated incidence") +
  scale_color_brewer(palette = "Set2") +
  theme(legend.position = "none")
```
