_---
title: "PLAN for Chapter 1. Modelology for fitting ODE epidemic models"
subtitle: "This RMarkdown is a template as initially outlined in the progress report. Each task will be outlined and then followed by a code block to be completed."
output:
html_notebook:
number_sections: true
---

**This should be similar to `Chapter_01.Rmd` but with the experiment data structure improved to take advantage of more cores than the number of scenarios.**

# Introduction

When calibrating their vivax population transmission models, White and Champagne lack longitudinal time series data and are forced to make equilibrium assumptions at a point in time (initially from Griffin from his 2014 P.f model). Here, derivatives are set to zero and the equations solved for unknowns. In the case of Champagne, the transmission rate $\lambda$ so that force of infection is $\lambda$ multiplied by the sum of infectious compartments $\sum{I}$.

This assumption does not hold for areas with moderate to strong seasonality and long-term trends. This document demonstrates the standard implementation of dynamic ODE parameter-fitting methods applied to the White/Champagne style model and extends the form of the model (and its corresponding parameter estimation routine) to provide increasingly flexible relaxations to the original uses of the White/Champagne model. The end result will be a model and fitting routine that is flexible enough to reflect vivax epidemics in non-stationary regions, unlike the original implementation.

[...]

# Models

Stan and posterior sampling will be used for all analyses.

```{r setup}
library(R.utils)
library(tidyverse)
library(rstan)
library(rstansim) # devtools::install_github("ewan-keith/rstansim")
library(parallel)
library(patchwork)
library(pbmcapply)
library(pbapply)
library(memoise)
library(RColorBrewer)
source("../R/constants.R")
source("../R/load_functions.R")

model_champagne2022_seasonal_ext = "../stan/champagne2022_seasonal_ext.stan"
stan_model_champagne2022_seasonal_ext = stan_model(model_champagne2022_seasonal_ext)
model_champagne2022 = "../stan/champagne2022.stan"
stan_model_champagne2022 = stan_model(model_champagne2022)

n_cores = parallelly::availableCores()
options(mc.cores = n_cores)
message("Running on ", n_cores, " cores")
rstan_options(auto_write = TRUE)

n_years = 5
n_iter = 100 # should be at least 500
n_chains = 4
n_repetitions = 1 # how many times to duplicate each scenario
cores_per_sampler = n_chains # 1 for mclapply, or n_chains if not running lots of scenarios
limit_runs = Inf # set to a finite number for testing, or Inf to run all
timelimit_per_run = 60*60 * 5
n_traces = 200 # limit for plotting sample trajectories

methods = methods %>%
  head(4)

param_colors = brewer.pal(4, "Set2")
names(param_colors) = c("lambda", "phi", "eps", "kappa")
```

## Model implementation

We begin with extending Champagne's 2022 model for tropical vivax to include seasonality.

First we implement and verify the parameter recovery ability of Stan with the Champagne model as published in 2022.

```{r}
# perform simulation study
dt = years/annual_subdivisions
t0 = -50*years
t = seq_len(n_years*annual_subdivisions) * dt
n_times = length(t)
N = 1.8e6/100 # population size

#initial conditions
I_init = 0.01
y0 = c(Il=0, I0=I_init, Sl=0, S0=1-I_init, CumulativeInfections=0)

# constants for Stan
data_consts = list(n_times = n_times+1,
                   y0 = y0,
                   t0 = t0,
                   ts = seq_len(n_times+1) * dt,
                   N = N,
                   cases = rep(99999, n_times+1),
                   r = 1./60, # r
                   gammal = 1./223, # gammal
                   f = 1./72, # f
                   alpha = 0.8, # alpha,
                   beta = 0.429/0.8, # beta, Nekkab 2021
                   delta = 0,
                   eps = 0,
                   kappa = 1,
                   phase = 0
)
```

## Parameter recovery on real data

We test whether parameter recovery works on real datasets. We will find data from a variety of settings (e.g., transmission levels, remoteness, strains) to demonstrate generalisability.

Data:

- Hainan data (tropical China) but this has been difficult to acquire.
We will not be investigating Chinese Yunnan (southern mountainous) or Henan (central temperate) data which we do have because the Yunnan strain is not known to be tropical, and the Henan data is temperate (our current temperate model does not align with this data convincingly).
- Brazilian 'integrated data set', available per county or municipality and very detailed.

```{r}
library(MalariaData)

integrated_dataset %>%
  count(notification.date) %>%
  ggplot(aes(x = notification.date, y = n)) +
  geom_col() +
  labs(title = "Lab-confirmed P. vivax cases in Brazil",
       x = "Month",
       y = "Count")

ggsave("../plots/brazil_cases.png", width=8, height=2)
```

```{r}
rondonia.all = integrated_dataset %>%
  filter(exam.result == "vivax",
         infection.state == "RO") %>%
  count(notification.year, notification.date, name = "cases") %>%
  mutate(include = ifelse(notification.year >= 2015 & notification.year <= 2018, "grey20", "grey")) %>%
  as_tibble()

ggplot(rondonia.all, aes(x = notification.date, y=cases, fill=include)) +
  annotate("rect", xmin=as.Date("2015-01-01"), xmax=as.Date("2018-12-31"), ymin=-Inf, ymax=Inf,
           fill = "blue", alpha = 0.15) +
  geom_col() +
  coord_cartesian(ylim = c(0, NA)) +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  scale_fill_identity() +
  labs(title = "Selected monthly P. vivax cases infected in Rondonia",
       x = "Notification date",
       y = "Cases")

ggsave("../plots/rondonia_selected.png", width=8, height=2)

rondonia = rondonia.all %>%
  filter(include == "grey20")

date_0 = min(rondonia$notification.date)
```

```{r}
source("../R/functions/decompose_data.R")
rondonia_decomp = rondonia %>%
  filter(notification.year >= 2015, notification.year <= 2018) %>%
  dplyr::rename(Date = notification.date,
                Cases = cases) %>%
  mutate(logCases = log(Cases)) %>%
  decompose_data()

rondonia_decomp %>%
  filter(name %in% c("Cases", "trend", "seasonal", "remainder")) %>%
  ggplot(aes(x = Date, y = value, color = value)) +
  geom_line() +
  coord_cartesian(ylim = c(0, NA)) +
  facet_wrap(vars(name), scales="free_y")
```

Create data for the full model

```{r}
data_ro = data_consts

data_ro$cases = rondonia$cases
data_ro$ts = seq_len(nrow(rondonia)) * years / annual_subdivisions
data_ro$n_times = nrow(rondonia)
data_ro$phase = 365.25*0.5
```

Create data for the nonseasonal model

```{r}
rondonia_agg = integrated_dataset %>%
  filter(exam.result == "vivax",
         infection.state == "RO",
         notification.year >= 2014) %>%
  mutate(notification.date = as.Date(paste(notification.year, "01", "01", sep="-")) + years(1) - days(1)) %>%
  count(notification.date, name = "cases")

# Aggregate data to a yearly level
data_ro_agg = data_consts

data_ro_agg$cases = rondonia_agg$cases
data_ro_agg$ts = seq_len(nrow(rondonia_agg)) * years
data_ro_agg$n_times = nrow(rondonia_agg)
data_ro_agg$y0 = data_consts$y0
data_ro_agg$t0 = data_consts$t0

tibble(ts = data_ro_agg$ts,
       cases = data_ro_agg$cases) %>%
  ggplot(aes(x = ts, y = cases)) +
  geom_line() +
  coord_cartesian(ylim = c(0, NA))
```

Perform MCMC sampling for both models

```{r}
samp_ext = sampling(stan_model_champagne2022_seasonal_ext,
                    data = data_ro,
                    iter = n_iter,
                    chains = n_chains,
                    # init = rep(list(.theta_init), n_chains), # Start from MLE solution
                    cores = n_cores)

samp_ext_nonseasonal = sampling(stan_model_champagne2022,
                                data = data_ro_agg,
                                iter = n_iter,
                                chains = n_chains,
                                # init = rep(list(.theta_init), n_chains), # Start from MLE solution
                                cores = n_cores)
```


```{r}
# inspect parameter posteriors
posterior_seasonal = rstan::extract(samp_ext, c("lambda", "phi_inv", "eps", "kappa")) %>%
  as_tibble()

p_seasonal_posterior = posterior_seasonal %>%
  mutate(phi = 1/phi_inv,
         phi_inv = NULL) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  ggplot(aes(x = value, fill=parameter, color=parameter)) +
  geom_density(alpha=0.5) +
  scale_fill_manual(values = param_colors, drop=F) +
  scale_color_manual(values = param_colors, drop=F) +
  coord_cartesian(xlim = c(0, NA)) +
  facet_wrap(vars(parameter), scales="free") +
  labs(subtitle = "Seasonal model")

# ggsave("plots/rondonia_seasonal_posterior.png", width=8, height=4)


# Plot credible intervals
incidence = rstan::extract(samp_ext, "incidence")[[1]]
sample_ix = sample(seq_len(dim(incidence)[1]), n_traces, replace=T)
incidence_sample_nonseasonal = as_tibble(t(incidence[sample_ix,])) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
  group_by(j) %>%
  mutate(ts = data_ro$ts[j],
         lower = quantile(incidence, 0.025),
         upper = quantile(incidence, 0.975),
         legend = "95% prediction interval")

sim_cases = rstan::extract(samp_ext, "sim_cases")[[1]]
sim_cases_sample_nonseasonal = as_tibble(t(sim_cases[sample_ix,])) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "sim_cases") %>%
  group_by(j) %>%
  mutate(ts = data_ro$ts[j],
         lower = quantile(sim_cases, 0.025),
         upper = quantile(sim_cases, 0.975),
         legend = "95% prediction interval") %>%
  distinct(j, .keep_all=T)

p_seasonal_ci = ggplot(mapping = aes(x=ts + date_0)) +
  geom_ribbon(data = sim_cases_sample_nonseasonal,
              aes(ymin=lower, ymax=upper, fill="95% case prediction interval"),
              alpha = 0.5) +
  geom_ribbon(data = incidence_sample_nonseasonal,
              aes(ymin=lower, ymax=upper, fill="95% mean incidence interval"),
              alpha = 0.5) +
  geom_point(data = tibble(ts = data_ro$ts, cases = data_ro$cases),
             aes(y = cases, group = NULL, color="Incidence data")) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_fill_manual("", 
                    breaks = c("95% case prediction interval", "95% mean incidence interval", "Incidence data"),
                    values = c("steelblue", "tomato", "black")) +
  scale_colour_manual("", 
                      breaks = c("95% case prediction interval", "95% mean incidence interval", "Incidence data"),
                      values = c("steelblue", "tomato", "black")) +
  labs(subtitle = "Seasonal model, monthly data",
       x = "Year",
       y = "Annual incidence")

# ggsave("plots/rondonia_seasonal_ci.png", width=8, height=4)
```

And, non-seasonal

```{r}
# inspect parameter posteriors
posterior_nonseasonal = rstan::extract(samp_ext_nonseasonal, c("lambda", "phi_inv")) %>%
  as_tibble()

# Plot posterior densities
p_ns_posterior = posterior_nonseasonal %>%
  mutate(phi = 1/phi_inv,
         phi_inv = NULL) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  ggplot(aes(x = value, fill=parameter, color=parameter)) +
  geom_density(alpha=0.5) +
  scale_fill_manual(values = param_colors, drop=F) +
  scale_color_manual(values = param_colors, drop=F) +
  coord_cartesian(xlim = c(0, NA)) +
  facet_wrap(vars(parameter), scales="free") +
  labs(subtitle = "Non-seasonal model")

# ggsave("plots/rondonia_ns_posterior.png", width=8, height=4)

# Plot traces
incidence = rstan::extract(samp_ext_nonseasonal, "incidence")[[1]]
sample_ix = sample(seq_len(dim(incidence)[1]), n_traces, replace=T)
incidence_sample_nonseasonal = as_tibble(t(incidence[sample_ix,])) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
  mutate(ts = data_ro_agg$ts[j],
         lower = quantile(incidence, 0.025),
         upper = quantile(incidence, 0.975),
         legend = "95% prediction interval")

sim_cases = rstan::extract(samp_ext_nonseasonal, "sim_cases")[[1]]
sim_cases_sample_nonseasonal = as_tibble(t(sim_cases[sample_ix,])) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "sim_cases") %>%
  mutate(ts = data_ro_agg$ts[j],
         lower = quantile(sim_cases, 0.025),
         upper = quantile(sim_cases, 0.975),
         legend = "95% prediction interval") %>%
  distinct(ts, .keep_all=T)

p_ns_ci = ggplot(mapping = aes(x=ts + date_0)) +
  geom_ribbon(data = sim_cases_sample_nonseasonal,
              aes(ymin=lower, ymax=upper, fill="95% case prediction interval"),
              alpha = 0.5) +
  geom_ribbon(data = incidence_sample_nonseasonal,
              aes(ymin=lower, ymax=upper, fill="95% mean incidence interval"),
              alpha = 0.5) +
  geom_point(data = tibble(ts = data_ro_agg$ts, cases = data_ro_agg$cases),
             aes(y = cases, group = NULL, color="Incidence data")) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_fill_manual("", 
                    breaks = c("95% case prediction interval", "95% mean incidence interval", "Incidence data"),
                    values = c("steelblue", "tomato", "black")) +
  scale_colour_manual("", 
                      breaks = c("95% case prediction interval", "95% mean incidence interval", "Incidence data"),
                      values = c("steelblue", "tomato", "black")) +
  labs(subtitle = "Non-seasonal model, annual data",
       x = "Year",
       y = "Annual incidence")

# ggsave("plots/rondonia_ns_ci.png", width=8, height=4)
```

Generate subplots

```{r}
p_seasonal_ci + p_ns_ci + plot_layout(guides = "collect") +
  plot_annotation(title = "Posterior model fits",
                  tag_levels = "A")
ggsave("plots/rondonia_ci.png", width=8, height=4)

p_seasonal_posterior + theme(legend.position="none") + p_ns_posterior + theme(legend.position="none") +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Posterior parameter estimates",
                  subtitle = "P. vivax transmission in Rondonia, 2015-2018",
                  tag_levels = "A")

ggsave("plots/rondonia_posterior.png", width=8, height=4)
```

Examine the estimated seasonal pattern

```{r}
omega = function(t, eps, kappa, phase=365.25/2) {
  return(eps + (1-eps)*pi/beta(0.5, kappa+0.5)*((1+sin(2*pi*(t - phase)/365.25))/2)^kappa)
}

omega_params = rstan::extract(samp_ext, c("eps", "kappa")) %>%
  as_tibble() %>%
  mutate(i = row_number())

tt = seq(0, 365.25, length.out=101)
omega_ts = lapply(seq_len(nrow(omega_params)), function(i) {
  omega(tt, omega_params$eps[i], omega_params$kappa[i])
})
omega_params$omega = omega_ts
omega_params$t = rep(list(tt), nrow(omega_params))
omega_params %>%
  unnest(c(t, omega)) %>%
  ggplot(aes(x = t/years*annual_subdivisions, y=omega, group=i)) +
  geom_line(alpha=0.05, color="grey10") +
  geom_hline(yintercept = 1, linetype="dashed") +
  coord_cartesian(ylim = c(0, NA)) +
  scale_x_continuous(breaks = 0:annual_subdivisions, expand = c(0,0)) +
  labs(title = "Posterior estimate of ω(t) for Rondonia P. vivax 2014-2018",
       x = "t (months)",
       y = "ω(t)")

ggsave("plots/rondonia_omega.png", width=8, height=4)
```

What if we used the steady-state fit to simulate posteriors in the full model?

```{r}
n_samples = 500
composite_list = pbmclapply(seq_len(n_samples), function(i) {
  theta_composite = list(
    lambda = sample(posterior_nonseasonal$lambda, 1),
    phi_inv = 0.1, # This doesn't matter if not sampling cases sample(posterior_seasonal$phi_inv, 1)
    eps = sample(posterior_seasonal$eps, 1),
    kappa = sample(posterior_seasonal$kappa, 1)
  )
  
  my_simulate_data(
    file = model_champagne2022_seasonal_ext,
    path = "sim_data",
    data_name = "rondonia_fit",
    input_data = data_ro,
    param_values = theta_composite,
    vars = c("ts", "incidence", "sim_cases")
  )
})
composite_list[sapply(composite_list, function(x) {"try-error" %in% class(x)})] <- NULL

incidence = rstan::extract(samp_ext, "incidence")[[1]]
ix = sample(seq_len(dim(incidence)[1]), n_traces, replace=T)
ts_sample = as_tibble(t(incidence[ix,])) %>%
  mutate(j = row_number()) %>%
  pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
  mutate(ts = data_ro$ts[j])

posterior_comparison = composite_list %>%
  bind_rows(.id = "trace") %>%
  mutate(method = "Nonseasonal calibration\nwith seasonality added") %>%
  bind_rows(ts_sample %>% mutate(method = "Seasonal model calibration", j=NULL)) %>%
  as_tibble()

posterior_comparison %>%
  ggplot(aes(x = ts+date_0, y = incidence)) +
  geom_line(aes(color=method, group=trace), alpha=0.1) +
  geom_point(data = tibble(ts = data_ro$ts, cases = data_ro$cases, color="Incidence data"), aes(y=cases)) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_colour_manual(NULL, 
                      breaks = c("Incidence data", "Seasonal model calibration", "Nonseasonal calibration\nwith seasonality added"),
                      values = c("black", "tomato", "steelblue")) +
  guides(color = guide_legend(override.aes = list(alpha = 1))) +
  labs(title = "Calibrated using the full seasonal model or the steady-state approximation",
       color = "Calibration routine",
       x = "Year",
       y = "Cases / mean incidence")

ggsave("plots/rondonia_composite.png", width=8, height=4)
```

How does this fitted seasonality compare against Martens (Malaria Atlas Project)

```{r}
library(sf)
rondonia_file = "~/Documents/MalariaData/data-raw/brazil/rondonia/data_EPSG_4326/11MIE250GC_SIR.shp" # https://purl.stanford.edu/df159kw0955

rondonia_geom = st_read(rondonia_file) %>%
  st_union() %>%
  st_as_sf() %>%
  mutate(Name = "Rondonia")
```


```{r}
map_folder = "~/Documents/MalariaData/data-raw/TSI-Martens2-Pv.2000-2022.Data.5km.Data"

# map = read_map(map_folder)
# map_sites = extract_layers(map, sites, "month", FALSE)
tifs = tibble(path = list.files(map_folder, full.names=T)) %>%
  mutate(filename = basename(path),
         year = str_extract(filename, "(1|2)[0-9]{3}") %>% as.numeric()) %>%
  filter(year >= 2014, year <= 2018)

x = terra::rast(tifs$path)
```

```{r}
rondonia_map = terra::extract(x, rondonia_geom) %>%
  pivot_longer(-ID, names_to="layer") %>%
  group_by(ID, layer) %>%
  summarise(value = mean(value, na.rm=T),
            .groups = "drop") %>%
  mutate(Name = rondonia_geom$Name[ID],
         year = layer %>% str_extract("[0-9]{4}") %>% as.numeric(),
         month = layer %>% str_extract("\\.[0-9]{2}\\.") %>% str_extract("[0-9]+") %>% as.numeric(),
         yearmonth = year*12 + month,
         date = as.Date(paste(year, month, "15", sep="-")))

rondonia_map_mean = rondonia_map %>%
  group_by(month) %>%
  summarise(value = mean(value))

ggplot(rondonia_map, aes(x=month, y=value, color=year, group=year)) +
  geom_line() +
  geom_line(data=rondonia_map_mean, aes(color=NULL, group=NULL), linetype="dashed") +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(limits = 0:1)
```

Generalise to a piecewise function
What data are we using?

```{r}
rondonia.nonstationary = integrated_dataset %>%
  filter(exam.result == "vivax",
         infection.state == "RO") %>%
  mutate(notification.date = as.Date(paste(notification.year, notification.month, "01", sep="-")) + months(1) - days(1)) %>%
  count(notification.year, notification.date, name = "cases") %>%
  mutate(include = ifelse(notification.year >= 2009 & notification.year <= 2018, "grey20", "grey")) %>%
  as_tibble()

p1 = ggplot(rondonia.nonstationary, aes(x = notification.date, y=cases, fill=include)) +
  annotate("rect", xmin=as.Date("2009-01-01"), xmax=as.Date("2018-12-31"), ymin=-Inf, ymax=Inf,
           fill = "blue", alpha = 0.15) +
  geom_col() +
  coord_cartesian(ylim = c(0, NA)) +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  scale_fill_identity() +
  labs(title = "Selected monthly P. vivax cases infected in Rondonia",
       x = "Notification date",
       y = "Cases")

library(MalariaData)
hainan.nonstationary = hainan %>%
  mutate(notification.date = as.Date(paste(Year, Month, "01", sep="-")) + months(1) - days(1),
         cases = `Vivax Cases`,
         include = "grey20")

t0_nonstationary = tibble(
  Region = c("Rondonia", "Hainan"),
  notification.date = as.Date(c("2011-01-01", "2004-01-01"))
)

plotdata_nonstationary = bind_rows(
  Rondonia = rondonia.nonstationary,
  Hainan = hainan.nonstationary,
  .id = "Region"
)
ggplot(plotdata_nonstationary, aes(x = notification.date, y=cases, fill=include)) +
  # annotate("rect",
  #          xmin=min(hainan.nonstationary$notification.date),
  #          xmax=max(hainan.nonstationary$notification.date),
  #          ymin=-Inf, ymax=Inf,
  #          fill = "blue", alpha = 0.15) +
  geom_col() +
  geom_vline(data=t0_nonstationary, aes(xintercept = notification.date), color="tomato") +
  ggrepel::geom_label_repel(data=t0_nonstationary, aes(x = notification.date, y=0, fill=NULL, label="t0"), color="tomato") +
  facet_wrap("Region", ncol=1, scales="free_y") +
  coord_cartesian(ylim = c(0, NA)) +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  scale_fill_identity() +
  labs(title = "Nonstationary trends in monthly P. vivax cases",
       x = "Notification date",
       y = "Cases")


ggsave("../plots/nonstationary_selected.png", width=8, height=4)
```


End of script.

```{r}
print("Done.")
```
