_---
title: "PLAN for Chapter 1. Methodology for fitting ODE epidemic models"
subtitle: "This RMarkdown is a template as initially outlined in the progress report. Each task will be outlined and then followed by a code block to be completed."
output:
html_notebook:
number_sections: true
---


Do the basic comparison between seasonal and non-seasonal model fitting

```{r setup}
library(R.utils)
library(tidyverse)
library(rstan)
library(rstansim) # devtools::install_github("ewan-keith/rstansim")
library(parallel)
library(patchwork)
library(ggridges)
library(pbmcapply)
library(pbapply)
library(memoise)
library(RColorBrewer)
source("../R/constants.R")
source("../R/load_functions.R")

n_cores = parallelly::availableCores()
# n_cores = 6
options(mc.cores = n_cores)
message("Running on ", n_cores, " cores")
rstan_options(auto_write = TRUE)

n_iter = 500 # should be at least 500
n_chains = 1
n_repetitions = 5 # how many times to duplicate each scenario
cores_per_sampler = n_chains # 1 for mclapply, or n_chains if not running lots of scenarios
limit_runs = Inf # set to a finite number for testing, or Inf to run all
timelimit_per_run = 3*hr
n_traces = 100 # limit for plotting sample trajectories
n_traces = min(n_traces, n_iter)
```

## Model implementation

Generate some data

```{r}
#initial conditions
I_init = 0.01
y0 = c(Il=0, I0=I_init, Sl=0, S0=1-I_init, CumulativeInfections=0)
n_years = 10

# constants for Stan
data_consts = list(y0 = y0,
                   t0 = -100*years,
                   N = 100000, # population size
                   r = 1./60, # r
                   gammal = 1./223, # gammal
                   f = 1./72, # f
                   alpha = 0.21, # alpha
                   beta = 0.66, # beta
                   rho = 0.21, # rho
                   delta = 0
)
```

Simulate some dummy data

```{r}
n_times = annual_subdivisions * n_years
data = data_init(data_consts,
                 ts = seq_len(n_times)*years/annual_subdivisions,
                 cases = rep(1, n_times),
                 n_times = n_times)
params = list(lambda=0.01, phi_inv=0.1, eps=0.001, kappa=1, phase=130)

sim = my_simulate_data_list(
  file = "../stan/champagne2022_seasonal_ext_2.stan",
  path = "sim_data",
  data_name = "seasonal",
  input_data = data,
  param_values = params,
  vars = c("ts", "incidence", "sim_cases")
)

synth_df = with(sim, tibble(
  ts = ts,
  incidence = incidence,
  cases = cases
))
```

Insert the new seasonal data into a data list and run the seasonal model on it.

```{r}
# Edit data with generated values
data_sim = data_init(data_consts,
                     n_times = nrow(synth_df),
                     ts = synth_df$ts,
                     cases = synth_df$cases)

data_sim_agg = aggregate_data(data_sim)
synth_df_agg = with(data_sim_agg, tibble(ts = ts, cases = cases)) 

with(data_sim, tibble(ts = ts/years, cases = cases)) %>%
  ggplot(aes(x=ts, y=cases)) +
  geom_col() +
  scale_x_continuous(breaks = 0:100)
with(data_sim_agg, tibble(ts = ts/years, cases = cases)) %>%
  ggplot(aes(x=ts, y=cases)) +
  geom_col() +
  scale_x_continuous(breaks = 0:100)
```

```{r, eval=F}
fit_seasonal = sampling(stan_model("../stan/champagne2022_seasonal_ext_2.stan"),
                        data = data_sim,
                        iter = n_iter,
                        init = list(params),
                        chains = n_chains)

fit_nonseasonal = sampling(stan_model("../stan/champagne2022_poisson.stan"),
                           data = data_sim_agg,
                           iter = n_iter,
                           init = list(params),
                           chains = n_chains)
```

Manually inspect fits

```{r, eval=F}
post_seasonal = rstan::extract(fit_seasonal, c("lambda", "eps", "kappa", "phase")) %>%
  as_tibble() %>%
  pivot_longer(everything())

ggplot(post_seasonal, aes(x=value, color=name, fill=name)) +
  geom_density() +
  facet_wrap(vars(name), scales="free")

incidence_seasonal = extract_incidence(fit_seasonal, data_sim$ts)
ggplot(incidence_seasonal, aes(x=ts, y=incidence, group=trace)) +
  geom_line(alpha = 0.05) +
  geom_point(data = synth_df, aes(y=cases, group=NULL)) +
  coord_cartesian(ylim = c(0, NA))

post_nonseasonal = rstan::extract(fit_nonseasonal, c("lambda")) %>%
  as_tibble() %>%
  pivot_longer(everything())

ggplot(post_nonseasonal, aes(x=value, color=name, fill=name)) +
  geom_density() +
  facet_wrap(vars(name), scales="free")

incidence_nonseasonal = extract_incidence(fit_nonseasonal, data_sim_agg$ts)
ggplot(incidence_nonseasonal, aes(x=ts, y=incidence, group=trace)) +
  geom_line(alpha = 0.05) +
  geom_point(data = synth_df_agg, aes(y=cases, group=NULL)) +
  coord_cartesian(ylim = c(0, NA))
```

# Can we make a variety of scenarios?
Define a function to simulate cases for a scenario

```{r}
simulate_cases = function(data_consts, lambda, phi_inv, eps, kappa, phase) {
  n_times = annual_subdivisions * n_years
  data = data_init(data_consts,
                   cases = rep(1, n_times),
                   ts = seq_len(n_times) * years / annual_subdivisions,
                   n_times = n_times)
  params = list(lambda=lambda, phi_inv=phi_inv, eps=eps, kappa=kappa, phase=phase)
  synth_df = my_simulate_data(
    file = "../stan/champagne2022_seasonal_ext_2.stan",
    path = "sim_data",
    data_name = "seasonal",
    input_data = data,
    param_values = params,
    vars = c("ts", "incidence", "sim_cases")
  )
}
```

Create scenarios

```{r}
# Define the scenarios
range_lambda = seq(0.01, 0.012, length.out=3)
range_eps = seq(1e-12, 1-1e-12, length.out=3)
scenarios = expand_grid(
  lambda = range_lambda,
  phi_inv = 0.1,
  eps = range_eps,
  kappa = 1,
  phase = 130,
  rep = seq_len(n_repetitions)
) %>%
  mutate(ID = row_number(), .before=0)

# Simulate cases for each scenario
cases_scenarios = pbmclapply(seq_len(nrow(scenarios)), function(i) {
  lambda = scenarios$lambda[i]
  phi_inv = scenarios$phi_inv[i]
  eps = scenarios$eps[i]
  kappa = scenarios$kappa[i]
  phase = scenarios$phase[i]
  simulate_cases(data_consts, lambda, phi_inv, eps, kappa, phase)
})

# Attach simulation results to the scenarios dataframe
scenarios$ts = lapply(cases_scenarios, function(x) {x$ts})
scenarios$cases = lapply(cases_scenarios, function(x) {x$cases})

# Duplicate for each method
scenarios_long = scenarios %>%
  cross_join(tibble(method = c("../stan/champagne2022_seasonal_ext_2.stan",
                               "../stan/champagne2022_poisson.stan")))
# Replace data with aggregate data depending on the method
for (i in seq_len(nrow(scenarios_long))) {
  if (scenarios_long$method[i] == "../stan/champagne2022_poisson.stan") {
    scenarios_long$ts[[i]] = aggregate_time(scenarios_long$ts[[i]])
    scenarios_long$cases[[i]] = aggregate_cases(scenarios_long$cases[[i]])
  }
}
```

What do our synthetic data look like?

```{r}
scenarios_long %>%
  mutate(method = case_when(method %>% str_detect("seasonal") ~ "Seasonal",
                            TRUE ~ "Nonseasonal")) %>%
  # distinct(scenario_ID, .keep_all=T) %>%
  unnest(cols = c("ts", "cases")) %>%
  ggplot(aes(x = ts/years, y = cases, color=method, group=interaction(method, rep))) +
  geom_line(alpha = 0.6) +
  scale_x_continuous(breaks = 0:n_years) +
  scale_y_log10(labels = label_auto2) +
  facet_grid(rows = vars(lambda),
             cols = vars(eps),
             labeller = plot_labeller) +
  labs(title = "Input data for each scenario",
       subtitle = "Seasonal scenario in cases per month; nonseasonal scenario per year",
       x = "Year",
       y = "Cases",
       color = "Method")
ggsave("../plots/synth_data.png", width=8, height=4)
```

Define a function to perform the model fit, then apply it to each scenario

```{r}
# Define function that takes simulated data
my_sampling = function(method, ts, cases, true_lambda=NULL) {
  data = data_init(data_consts, n_times = length(ts), ts = ts, cases = cases)
  if (method == "../stan/champagne2022_poisson.stan") {
    result = sampling(stan_model(method),
                      data = data,
                      iter = n_iter,
                      init = list(list(lambda = true_lambda)),
                      chains = 1)
  } else {
    result = sampling(stan_model(method),
                      data = data,
                      iter = n_iter,
                      chains = 1)
  }
  result
}

# Apply function to simulated data
tictoc::tic("Running all scenarios")
fits_scenarios = pbmclapply(seq_len(nrow(scenarios_long)), function(i) {
  with_timeout(
    my_sampling(scenarios_long$method[i], scenarios_long$ts[[i]], scenarios_long$cases[[i]]),
    cpu = timelimit_per_run)
})
tictoc::toc()

# Redo for just nonseasonal runs
nonseasonal_ix = which(scenarios_long$method == "../stan/champagne2022_poisson.stan")
new_fits_scenarios = pbmclapply(nonseasonal_ix, function(i) {
  with_timeout(
    my_sampling(scenarios_long$method[i], scenarios_long$ts[[i]], scenarios_long$cases[[i]], scenarios_long$lambda[i]),
    cpu = timelimit_per_run)
})
fits_scenarios[nonseasonal_ix] = new_fits_scenarios
```

How many didn't time out?

```{r}
sapply(fits_scenarios, is.null)
sapply(fits_scenarios, function(x) {x@mode})
```

```{r}
# Extract parameters into dataframe
estimates_scenarios = lapply(seq_len(length(fits_scenarios)), function(i) {
  fit = fits_scenarios[[i]]
  if (scenarios_long$method[i] == "../stan/champagne2022_poisson.stan") {
    posterior = rstan::extract(fit, c("lambda")) %>%
      as_tibble() %>%
      mutate(phi_inv = NA, eps = NA, kappa = NA, phase = NA)
  } else {
    posterior = rstan::extract(fit, c("lambda", "phi_inv", "eps", "kappa", "phase")) %>%
      as_tibble()
  }
  return(posterior)
})

# Attach re-estimated parameters
scenarios_long$est_lambda = lapply(estimates_scenarios, function(x) {x$lambda})
scenarios_long$est_phi = lapply(estimates_scenarios, function(x) {1/x$phi_inv})
scenarios_long$phi = 1/scenarios_long$phi_inv
scenarios_long$est_eps = lapply(estimates_scenarios, function(x) {x$eps})
scenarios_long$est_kappa = lapply(estimates_scenarios, function(x) {x$kappa})
scenarios_long$est_phase = lapply(estimates_scenarios, function(x) {x$phase})

# Attach posterior predictions
scenarios_long$est_incidence = lapply(seq_len(nrow(scenarios_long)), function(i) {
  x = fits_scenarios[[i]]
  if (x@mode == 2) {
    return(NULL)
  }
  scenario = scenarios_long[i,]
  method = scenario$method
  data_sim = data_init(data_consts, n_times = length(scenario$ts[[1]]), ts = scenario$ts[[1]], cases = scenario$cases[[1]])
  incidence = rstan::extract(x, "incidence")[[1]]
  if (n_traces < dim(incidence)[1]) {
    sample_ix = seq_len(dim(incidence)[1])
  } else {
    sample_ix = sample(seq_len(dim(incidence)[1]), n_traces, replace=T)
  }
  
  incidence_sample_nonseasonal = as_tibble(t(incidence[sample_ix,])) %>%
    mutate(j = row_number()) %>%
    pivot_longer(-j, names_to = "trace", values_to = "incidence") %>%
    drop_na(j) %>%
    group_by(j) %>%
    mutate(ts = data_sim$ts[j],
           lower = quantile(incidence, 0.025, na.rm=T),
           upper = quantile(incidence, 0.975, na.rm=T),
           legend = "95% prediction interval",
           rep = scenario$rep)
})
```

Diagnostic plots

Firstly, inspect the seasonal model.

```{r}
plot_data = scenarios_long %>%
  filter(method %>% str_detect("seasonal")) %>%
  select(-ts, -cases, -est_incidence) %>%
  pivot_longer(matches("^est")) %>%
  unnest(value) %>%
  drop_na(value) %>%
  mutate(name = name %>% str_remove("^est_"),
         method = ifelse(method %>% str_detect("seasonal"), "Seasonal", "Nonseasonal")) %>%
  group_by(ID, name) %>%
  mutate(lq = quantile(value, 0.01),
         uq = quantile(value, 0.99)) %>%
  filter(value < uq) %>%
  ungroup()

true_values = scenarios_long %>%
  distinct(lambda, phi, eps, kappa, phase) %>%
  mutate(lambda2 = lambda,
         eps2 = eps) %>%
  pivot_longer(-c(lambda2, eps2)) %>%
  dplyr::rename(lambda = lambda2,
                eps = eps2)

my_labeller <- function(variable, value){
  var = make_greek(variable)
  if (is.numeric(value)) {
    value = round(value, 6)
  }
  if (variable == "name") {
    return(make_greek(value))
  } else {
    return(paste(var, "=", value))
  }
}

ggplot(plot_data, aes(x = value, y = as.factor(lambda), color = lambda, fill = lambda, group = interaction(as.factor(lambda), as.factor(rep)))) +
  geom_density_ridges(alpha = 0.25) +
  # geom_boxplot(outliers = FALSE) +
  # geom_violin(alpha = 0.25) +
  # geom_density() +
  geom_point(data = true_values, aes(color=NULL, fill = NULL, group=NULL), shape=1) +
  # facet_wrap(vars(name), scales="free_y")
  facet_grid(rows = vars(eps), cols = vars(name), scales="free", labeller = labeller(name = ~ make_greek(as.character(.)),
                                                                                     eps = ~ paste(make_greek("eps"), "=", round(as.numeric(.), 5)),
                                                                                     .multi_line=F)) +
  labs(title = "Posterior parameter densities of the seasonal model",
       subtitle ="True values denoted by circles",
       x = "Value",
       y = paste("Scenario", make_greek("lambda")),
       color = paste("Scenario", make_greek("lambda")),
       fill = paste("Scenario", make_greek("lambda"))) +
  theme(legend.position = "none")

ggsave("../plots/basic_seasonal_posterior.png", width=8, height=4)
```

```{r}
plot_data = scenarios_long %>%
  select(-ts, -cases, -est_incidence) %>%
  pivot_longer(matches("^est")) %>%
  unnest(value) %>%
  drop_na(value) %>%
  mutate(name = name %>% str_remove("^est_"),
         method = ifelse(method %>% str_detect("seasonal"), "Seasonal", "Nonseasonal")) %>%
  group_by(name) %>%
  mutate(lq = quantile(value, 0.01),
         uq = quantile(value, 0.99)) %>%
  filter(value > lq,
         value < uq)

true_values = scenarios_long %>%
  distinct(lambda, phi, eps, kappa, phase) %>%
  mutate(lambda2 = as.factor(lambda),
         eps2 = as.factor(eps)) %>%
  pivot_longer(-c(lambda2, eps2)) %>%
  dplyr::rename(lambda = lambda2,
                eps = eps2)

posterior_ranges = tribble(
  ~name, ~min, ~max,
  "lambda", 0, 0.015,
  "phi", 0, 20,
  "eps", 0, 1,
  "kappa", 0, 10,
  "phase", 0, years
)
posterior_ranges_data = posterior_ranges %>%
  pivot_longer(-name, names_to="bound")

plot_data = plot_data %>%
  inner_join(posterior_ranges, by="name") %>%
  filter(value >= min, value <= max)

my_labeller <- function(variable, value){
  var = make_greek(variable)
  if (is.numeric(value)) {
    value = round(value, 6)
  }
  if (variable == "name") {
    return(make_greek(value))
  } else {
    return(paste(var, "=", value))
  }
}

ggplot(plot_data, aes(x = value, y = as.factor(lambda), color = method, fill = method, group = interaction(lambda, rep, method))) +
  geom_density_ridges(alpha = 0.25) +
  geom_point(data = posterior_ranges_data, aes(y = as.factor(0.01), color = NULL, fill = NULL, group = NULL), alpha=0) +
  geom_point(data = true_values, aes(color=NULL, fill = NULL, group=NULL), shape=1) +
  facet_wrap(vars(eps, make_greek(name)), scales="free", nrow=3, labeller = labeller(eps = ~ paste(make_greek("eps"), "=", round(as.numeric(.), 5)),
                                                                                     .multi_line=F)) +
  theme(legend.position = "right") +
  labs(title = "Synthetic parameter estimation comparison over repeated scenarios",
       x = "Value", y = NULL, color = "Method", fill = "Method")
ggsave("../plots/basic_posterior_comparison.png", width=8, height=4)
```

Lambda only

```{r}
plot_data_2 = plot_data %>%
  filter(name == "lambda") %>%
  mutate(eps = as.numeric(as.character(eps)))
true_values_2 = true_values %>%
  filter(name == "lambda") %>%
  mutate(eps = as.numeric(as.character(eps))) %>%
  left_join(tibble(method = c("Seasonal", "Nonseasonal")), by=character())

ggplot(plot_data_2, aes(x = value, y = method, color = method, fill = method, group = interaction(lambda, rep, method))) +
  geom_vline(data=true_values_2, aes(xintercept = value), linetype="dashed") +
  geom_density_ridges(alpha = 0.25) +
  scale_x_continuous(breaks = unique(true_values_2$value)) +
  # geom_point(data = true_values_2, aes(y = method, color=NULL, fill = NULL, group=NULL), shape=1) +
  facet_grid(rows = vars(eps), cols = vars(lambda), labeller = plot_labeller, scales = "free_y") +
  theme(legend.position = "none") +
  labs(title = "Synthetic parameter estimation comparison over repeated scenarios",
       x = "Value", y = NULL, color = "Method", fill = "Method")

ggsave("../plots/basic_method_comparison.png", width=8, height=4)
```

Plot incidence fits

Single trace

```{r}
scenarios_long$est_incidence[[2]] %>%
  ggplot(aes(x=ts, y=incidence, color=rep, group=interaction(trace, rep))) +
  geom_line()

```

```{r}
incidence_join = lapply(seq_len(nrow(scenarios_long)), function(i) {
  if (is.null(scenarios_long$est_incidence[[i]])){
    return(NULL)
  }
  est_incidence = scenarios_long$est_incidence[[i]] %>%
    mutate(ID = scenarios_long$ID[i],
           method = scenarios_long$method[i],
           lambda = scenarios_long$lambda[i],
           eps = scenarios_long$eps[i],
           rep = scenarios_long$rep[i])
}) %>%
  bind_rows() %>%
  ungroup() %>%
  select(-lower, -upper, -legend)

if (n_traces < length(unique(incidence_join$trace))) {
  trace_sample = sample(unique(incidence_join$trace), n_traces)
  incidence_join = incidence_join %>%
    filter(trace %in% trace_sample)
}

# For each scenario, find a rep where nothing failed
first_successful_scenario = incidence_join %>%
  group_by(lambda, eps, method) %>%
  slice(1) %>%
  select(lambda, eps, method, rep) %>%
  ungroup()
incidence_join = incidence_join %>%
  inner_join(first_successful_scenario, by=c("rep", "method", "lambda", "eps")) %>%
  mutate(method = ifelse(method %>% str_detect("poisson"), "Nonseasonal", "Seasonal")) %>%
  mutate(incidence = ifelse(method == "Nonseasonal", incidence/12, incidence))


scenarios_overlay = scenarios %>%
  unnest(c(ts, cases)) %>%
  inner_join(first_successful_scenario, by=c("rep", "lambda", "eps"))

ggplot(incidence_join, aes(x = ts/years, y = incidence, color=method, group = interaction(method, rep, trace))) +
  geom_point(data=scenarios_overlay, aes(y=cases, color=NULL, group=NULL), alpha=0.25, show.legend=F, size=1) +
  geom_line(alpha = 0.25) +
  facet_grid(rows = vars(lambda), cols = vars(eps), labeller = plot_labeller, scales="free_y") +
  scale_x_continuous(breaks = 0:n_years) +
  scale_y_log10(labels = label_auto2) +
  labs(title = "Mean estimated incidence rates for the first scenario repetition",
       subtitle = "Nonseasonal (annual) results scaled to per-month",
       x = "Year",
       y = "Incidence", color = "Method") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))

ggsave("../plots/basic_modelled_incidence.png", width=8, height=4)
```

Re-simulate data with the true seasonality

```{r}
valid_scenarios_long = scenarios_long %>%
  inner_join(first_successful_scenario)
resimulated = lapply(seq_len(nrow(valid_scenarios_long)), function(i) {
  if (is.null(valid_scenarios_long$est_incidence[[i]])){
    return(NULL)
  }
  pbmclapply(seq_len(8), function(j) {
    if (is.null(valid_scenarios_long$est_lambda[[i]])) {
      return(NULL)
    }
    lambda = sample(valid_scenarios_long$est_lambda[[i]], 1)
    phi_inv = valid_scenarios_long$phi_inv[i]
    eps = valid_scenarios_long$eps[i]
    kappa = valid_scenarios_long$kappa[i]
    phase = valid_scenarios_long$phase[i]
    x = simulate_cases(data_consts, lambda, phi_inv, eps, kappa, phase)
    if (is.data.frame(x)) {
      result = x %>%
      mutate(lambda = valid_scenarios_long$lambda[i],
             eps = eps,
             method = valid_scenarios_long$method[i],
             trace = j)
      return(result)
    } else {
      return(NULL)
    }
    
  }) %>%
    bind_rows()
})
resimulated_2 = resimulated %>%
  bind_rows() %>%
  mutate(method = ifelse(method %>% str_detect("poisson"), "Nonseasonal", "Seasonal"))

ggplot(resimulated_2, aes(x = ts/years, y = incidence, color=method, group = interaction(method, trace))) +
  geom_point(data=scenarios_overlay, aes(y=cases, color=NULL, group=NULL), alpha=0.25, show.legend=F, size=1) +
  geom_line(alpha = 0.25) +
  facet_grid(rows = vars(lambda), cols = vars(eps), labeller = plot_labeller, scales="free_y") +
  scale_x_continuous(breaks = 0:n_years) +
  scale_y_log10(labels = label_auto2) +
  labs(title = "Re-simulated incidence rates for the first scenario repetition",
       # subtitle = "Nonseasonal (annual) results scaled to per-month",
       x = "Year",
       y = "Incidence", color = "Method") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))

ggsave("../plots/basic_resimulation.png", width=8, height=4)
```