_---
title: "PLAN for Chapter 1. Methodology for fitting ODE epidemic models"
subtitle: "This RMarkdown is a template as initially outlined in the progress report. Each task will be outlined and then followed by a code block to be completed."
output:
html_notebook:
number_sections: true
---


This document is for fitting an epidemic model where the system begins out of equilibrium and gradually returns to equilibrium.

Method:

1. Lambda takes the form of a step function: lambda(t) = *λ* for t<0, = *λ + Δλ* for t >= 0
2. We find an approximate solution for *λ* and *X* for t=0 by assuming equilibrium for t<0 at data point Incidence(t=0)
3. We calibrate *Δλ* and/or other parameters by fitting to data for t>=0

```{r setup}
library(R.utils)
library(tidyverse)
library(rstan)
library(rstansim) # devtools::install_github("ewan-keith/rstansim")
library(parallel)
library(patchwork)
library(pbmcapply)
library(pbapply)
library(memoise)
library(RColorBrewer)
source("../R/constants.R")
source("../R/load_functions.R")

model_champagne2022_nonstationary = "../stan/champagne2022_nonstationary.stan"
stan_model_champagne2022_nonstationary = stan_model(model_champagne2022_nonstationary)

n_cores = parallelly::availableCores()
options(mc.cores = n_cores)
message("Running on ", n_cores, " cores")
rstan_options(auto_write = TRUE)

n_years = 5
n_iter = 500 # should be at least 500
n_chains = 4
n_repetitions = 1 # how many times to duplicate each scenario
cores_per_sampler = n_chains # 1 for mclapply, or n_chains if not running lots of scenarios
limit_runs = Inf # set to a finite number for testing, or Inf to run all
timelimit_per_run = 60*60 * 5
n_traces = 200 # limit for plotting sample trajectories

param_colors = brewer.pal(4, "Set2")
names(param_colors) = c("lambda", "phi", "eps", "kappa")
```

## Model implementation

Generate some data

```{r}
# perform simulation study
dt = years/annual_subdivisions
t0 = -1000*years
t = seq_len(n_years*annual_subdivisions) * dt
n_times = length(t) + 1

#initial conditions
I_init = 0.01
y0 = c(Il=0, I0=I_init, Sl=0, S0=1-I_init, CumulativeInfections=0)

# constants for Stan
data_consts = list(n_times = n_times,
                   y0 = y0,
                   t0 = t0,
                   ts = seq_len(n_times) * dt,
                   N = 100000, # population size
                   cases = rep(99999, n_times),
                   r = 1./60, # r
                   gammal = 1./223, # gammal
                   f = 1./72, # f
                   alpha = 0.21, # alpha
                   beta = 0.66, # beta
                   rho = 0.21, # rho
                   delta = 0,
                   phase = 0
)

n_data = annual_subdivisions * 100
data = data_consts
data$cases = rep(1, n_data)
data$ts = seq_len(n_data) * years / annual_subdivisions
data$n_times = n_data
data$phase = 365.25*0.5

params = list(lambda=0.01, phi_inv=0.1, eps=0.01, kappa=1, dlambda=-0.001)

sim = my_simulate_data(
  file = model_champagne2022_nonstationary,
  path = "sim_data",
  data_name = "nonstationary",
  input_data = data,
  param_values = params,
  vars = c("ts", "incidence", "sim_cases")
)

sim

ggplot(sim, aes(x=ts, y=cases)) +
  geom_line()
```

Insert the new seasonal data into a data list and run the seasonal model (MLE and posterior sampling) on it.

```{r, eval=F}
# Fit data
# Edit data with generated values
data = data_consts
data$n_times = nrow(sim)
data$ts = sim$ts
data$cases = sim$cases

fit = sampling(stan_model_champagne2022_nonstationary,
               data = data,
               iter = n_iter,
               chains = n_chains,
               seed = 0)

# pairs(fit_seasonal, pars=c("lambda", "phi_inv"))

smr_pred_seasonal <- with(sim, cbind(as.data.frame(
  summary(
    fit_seasonal,
    pars = "sim_cases",
    probs = c(0.05, 0.5, 0.95)
  )$summary),
  t=t[1:(n_times-1)], cases=cases[1:(n_times-1)])) %>%
  setNames(colnames({.}))

ggplot(smr_pred_seasonal, mapping = aes(x = t)) +
  geom_ribbon(aes(ymin = `5%`, ymax = `95%`), fill = c_posterior, alpha = 0.35) +
  geom_line(mapping = aes(y = `50%`), color = c_posterior) + 
  geom_point(mapping = aes(y = cases)) +
  labs(x = "Day", y = "Cases") +
  coord_cartesian(ylim = c(0, NA))
```
