_---
title: "PLAN for Chapter 1. Methodology for fitting ODE epidemic models"
subtitle: "This RMarkdown is a template as initially outlined in the progress report. Each task will be outlined and then followed by a code block to be completed."
output:
html_notebook:
number_sections: true
---

**This should be similar to `Chapter_01.Rmd` but with the experiment data structure improved to take advantage of more cores than the number of scenarios.**

# Introduction

When calibrating their vivax population transmission models, White and Champagne lack longitudinal time series data and are forced to make equilibrium assumptions at a point in time (initially from Griffin from his 2014 P.f model). Here, derivatives are set to zero and the equations solved for unknowns. In the case of Champagne, the transmission rate $\lambda$ so that force of infection is $\lambda$ multiplied by the sum of infectious compartments $\sum{I}$.

This assumption does not hold for areas with moderate to strong seasonality and long-term trends. This document demonstrates the standard implementation of dynamic ODE parameter-fitting methods applied to the White/Champagne style model and extends the form of the model (and its corresponding parameter estimation routine) to provide increasingly flexible relaxations to the original uses of the White/Champagne model. The end result will be a model and fitting routine that is flexible enough to reflect vivax epidemics in non-stationary regions, unlike the original implementation.

[...]

# Methods

Stan and posterior sampling will be used for all analyses.

```{r setup}
library(R.utils)
library(tidyverse)
library(rstan)
library(rstansim) # devtools::install_github("ewan-keith/rstansim")
library(parallel)
library(patchwork)
library(pbmcapply)
library(pbapply)
library(memoise)
source("../R/constants.R")
source("../R/methods.R")
n_cores = parallelly::availableCores()
options(mc.cores = n_cores)
message("Running on ", n_cores, " cores")
rstan_options(auto_write = TRUE)

# Store generated data here
cd = cachem::cache_disk("memoise")

n_years = 5
n_iter = 500 # should be at least 500
n_chains = 2
n_repetitions = 1 # how many times to duplicate each scenario
cores_per_sampler = 1 # set to n_chains if not running lots of scenarios
limit_runs = Inf # set to a finite number for testing, or Inf to run all
timelimit_per_run = 60*60
n_traces = 100 # limit for plotting sample trajectories

methods = methods %>%
  head(4)
```

## Model implementation

We begin with extending Champagne's 2022 model for tropical vivax to include seasonality.

First we implement and verify the parameter recovery ability of Stan with the Champagne model as published in 2022.

```{r}
# perform simulation study
dt = years/annual_subdivisions
t0 = -100*years
t = seq_len(n_years*annual_subdivisions) * dt
n_times = length(t)
N = 10000 # population size

#initial conditions
I_init = 0.01
y0 = c(Il=0, I0=I_init, Sl=0, S0=1-I_init, CumulativeInfections=0)

# constants for Stan
data_consts = list(n_times = n_times+1,
                   y0 = y0,
                   t0 = t0,
                   ts = seq_len(n_times+1) * dt,
                   N = N,
                   cases = rep(99999, n_times+1),
                   r = 1./60, # r
                   gammal = 1./223, # gammal
                   f = 1./72, # f
                   alpha = 0.615, # alpha, 0.437/0.713
                   beta = 0.713, # beta, Nekkab 2021
                   rho = 0.21, # rho
                   delta = 0,
                   eps = 0,
                   kappa = 1,
                   phase = 0
)
```

## Parameter recovery comparison on synthetic data

Under ideal conditions, model fitting using the time series should recovery the original parameters with reasonable accuracy, but the Champagne 2022 method will have some error term resulting from seasonality.

### Reasonable scenarios

We define reasonable scenarios that reflect the assumptions of the Champagne model except for seasonality. For example, a range of transmission intensities and treatment capabilities. Data will be taken from oscillating or steady-state periods after the long-term trend has been stabilised as we are only relaxing the non-seasonal assumption of Champagne 2022.

Scenarios include sequences of scenarios. For example, low-transmission tropical relapse with a range of magnitudes of seasonality (peak-trough ratio from 0% to 100%). This sequence would demonstrate how the magnitude of seasonality impacts the accuracy of parameter recovery.

```{r}
# ascertainment_rates = c(0.25, 0.5, 0.75, 1)
# radical_cure_rates = seq(0, 1, by=0.2)
ascertainment_rates = data_consts$alpha
radical_cure_rates = data_consts$beta
seasonality_ratio = seq(0, 1, length.out=3)
# radical_cure_rates = data_consts$beta
transmission_rates = seq(0.027, 0.03, length.out=3)
importation_rate = data_consts$delta # because constant importation makes less sense in seasonal transmission
population_size = data_consts$N
```

Expand scenarios into a grid and generate synthetic data.

```{r}
data_scenarios = expand_grid(
  ascertainment_rates,
  radical_cure_rates,
  transmission_rates,
  importation_rate,
  seasonality_ratio,
  population_size
) %>%
  mutate(scenario_ID = LETTERS[row_number()], .before=0)

.simulate_cases = function(alpha=0.5, beta=0.5, lambda=0.01, phi_inv=0.1, delta=0, eps=0, kappa=1, phase=0, N=100, index=NULL) {
  data = data_consts
  data$alpha = alpha
  data$beta = beta
  data$delta = delta
  data$phase = phase
  data$eps = eps
  data$N = N
  
  real_params = list(lambda=lambda, phi_inv=phi_inv, kappa=kappa)
  print(real_params)
  
  if (is.null(index)) {
    index = sample.int(999999999, 1)
  }
  synth_df = suppressMessages(my_simulate_data(
    file = model_champagne2022_seasonal,
    path = "sim_data",
    data_name = paste0("data_", index),
    input_data = data,
    param_values = real_params,
    vars = c("ts", "incidence", "sim_cases", "susceptible")
  ))
}
x = .simulate_cases(lambda=0.05)

# Add cases onto a dataframe of scenarios based on its parameter columns
simulate_cases = function(.scenarios) {
  cases_scenarios = pbmclapply(seq_len(nrow(.scenarios)), function(i) {
    dat = .scenarios[i,]
    x = .simulate_cases(alpha = dat$ascertainment_rates,
                        beta = dat$radical_cure_rates,
                        lambda = dat$transmission_rates,
                        phi_inv = 0.1,
                        delta = dat$importation_rate,
                        kappa = 1,
                        phase = 0,
                        eps = dat$seasonality_ratio,
                        N = dat$population_size,
                        index = i)
  })
  
  .scenarios$cases = lapply(cases_scenarios, function(x) {x$cases})
  .scenarios$ts = lapply(cases_scenarios, function(x) {x$ts})
  .scenarios
}

simulate_cases_memo = memoise(simulate_cases, cache=cd)

data_scenarios_sim = data_scenarios %>%
  slice(rep(1:n(), each = n_repetitions)) %>%
  group_by(scenario_ID) %>%
  mutate(rep = row_number()) %>%
  ungroup() %>%
  simulate_cases()
# simulate_cases_memo()

# Display scenarios
data_scenarios_sim %>%
  distinct(scenario_ID, .keep_all=T) %>%
  unnest(cols = c("ts", "cases")) %>%
  ggplot(aes(x = ts, y = cases, group=interaction(seasonality_ratio, transmission_rates))) +
  geom_point(alpha = 0.8) +
  scale_color_gradient(trans = "log", breaks=10^seq(-5, 5)) +
  scale_y_log10() +
  facet_grid(vars(seasonality_ratio), vars(transmission_rates))
```

### Comparison of parameter recovery

On each scenario or sequence, recover the parameters using Champagne's solution on annual data and the typical ODE method. Show the resulting errors to the true parameter value.

Below: For testing

Execute methods on each scenario

```{r}
data_scenarios_long = data_scenarios_sim %>%
  tidyr::crossing(methods) %>%
  head(limit_runs)

# Aggregate data for nonseasonal methods
for (i in seq_len(nrow(data_scenarios_long))) {
  if (data_scenarios_long$method[i] %>% str_detect("nonseasonal")) {
    data_scenarios_long$ts[[i]] = aggregate_time(data_scenarios_long$ts[[i]])
    data_scenarios_long$cases[[i]] = aggregate_time(data_scenarios_long$cases[[i]])
    data_scenarios_long$eps[[i]] = aggregate_eps(data_scenarios_long$eps[[i]])
  }
}
run_scenario_method_memo = memoise(run_scenario_method, cache=cd)

tictoc::tic()
fits_all = pbmclapply(seq_len(nrow(data_scenarios_long)), function(...) {
  run_scenario_method(..., force=T)
})

tictoc::toc()

workspace_filename = format(Sys.time(), "%Y%M%d %H%M%S.Rdata")
# save.image(workspace_filename)
```

Extract data from each result

```{r}
data_scenarios_long$est_lambda = lapply(fits_all, extract_lambda)
data_scenarios_long$est_phi_inv = lapply(fits_all, extract_phi_inv)
data_scenarios_long$time = lapply(fits_all, extract_runtime)
data_scenarios_long$est_incidence = lapply(fits_all, extract_incidence)

# check samples
sapply(data_scenarios_long$est_lambda, class)
sapply(data_scenarios_long$est_lambda, function(x) length(x))
```

Compare densities for individual scenarios between methods

```{r}
plot_data = data_scenarios_long %>%
  select(-ascertainment_rates, -radical_cure_rates, -importation_rate, -population_size) %>%
  mutate(rhat = calculate_rhat(est_lambda)) %>%
  unnest(c(est_lambda, rhat)) %>%
  # filter(rhat < 1.5) %>% # Get rid of instances that appeared to not converge
  drop_na(est_lambda) %>%
  group_by(scenario_ID, method, rep) %>%
  mutate(trace = row_number()) %>%
  ungroup()

for (i in seq_len(nrow(data_scenarios))) {
  id = data_scenarios$scenario_ID[i]
  this_scenario = data_scenarios_long %>%
    filter(scenario_ID == id)
  plot_cases_i = this_scenario %>%
    filter(method %>% str_detect("_sea")) %>%
    slice(1) %>%
    unnest(c(ts, cases)) %>%
    select(ts, cases)
  
  title = with(this_scenario, paste0(id, ": lambda=", transmission_rates[1], " eps=", seasonality_ratio[1]))
  
  # Case plot
  p1 = ggplot(plot_cases_i, aes(x = ts/years, y = cases)) +
    geom_point() +
    labs(x = "Years")
  
  # Density plot
  samples = this_scenario %>%
    select(est_lambda, method, rep) %>%
    unnest(est_lambda) %>%
    group_by(method, rep) %>%
    mutate(iteration = row_number())
  
  true_value = this_scenario$transmission_rates[1]
  xlim_buffer = 0.1
  limits = c(true_value * (1-xlim_buffer), true_value * (1+xlim_buffer))
  limits_2 = c(true_value * (1-2*xlim_buffer), true_value * (1+2*xlim_buffer))
  
  p2 = ggplot(samples, aes(x = est_lambda, fill = method, color = method, group=interaction(method, rep))) +
    geom_density(alpha=0.2) +
    geom_vline(aes(xintercept = true_value), linetype="dashed") +
    scale_colour_manual(values = comparison_colors) +
    scale_fill_manual(values = comparison_colors) +
    coord_cartesian(xlim = limits)
  
  p3 = ggplot(samples, aes(x = iteration, y = est_lambda, color = method, group=interaction(method, rep))) +
    geom_line(alpha=0.5) +
    scale_colour_manual(values = comparison_colors) +
    coord_cartesian(ylim = limits_2)
  
  print(p1 / p2 / p3 + plot_layout(guides = "collect")+
    plot_annotation(title = title)) 
}
```

Some of these chains appear to not mix.

```{r}
not_run = sapply(data_scenarios_long$est_lambda, is.null)
data_scenarios_long$time[not_run] = -10
data_scenarios_long %>%
  mutate(time = unlist(time)) %>%
  ggplot(aes(x = time, fill = method)) +
  geom_histogram(alpha = 0.7, binwidth=1) +
  geom_vline(xintercept = -0.5) +
  scale_fill_manual(values = comparison_colors) +
  labs(title = "Which methods took a long time?",
       subtitle = paste(sum(not_run), "scenarios terminated after timeout of", timelimit_per_run, "seconds are placed at -10")) +
  facet_grid(rows = vars(seasonality_ratio),
             cols = vars(transmission_rates))

rhat_threshold = 1.05
convergence = plot_data %>%
  group_by(scenario_ID, rep, method) %>%
  slice(1) %>%
  filter(rhat > rhat_threshold)

plot_data %>%
  group_by(scenario_ID, rep, method) %>%
  slice(1) %>%
  ggplot(aes(x = rhat, fill = method)) +
  geom_histogram(alpha = 0.7, binwidth=0.1) +
  scale_fill_manual(values = comparison_colors) +
  labs(title = "Did the runs converge?",
       subtitle = paste0(nrow(convergence), " runs had an rhat on lambda greater than ", rhat_threshold, " (max ", max(round(plot_data$rhat, 2)), ")")) +
  facet_grid(rows = vars(seasonality_ratio),
             cols = vars(transmission_rates))
```
What are the modelled incidence trends like?

```{r}
valid_ix = which(sapply(data_scenarios_long$est_lambda, class) == "array")
plot_data = lapply(valid_ix, function(i) {
  incidence_df = data_scenarios_long$est_incidence[[i]]
  if (is.null(incidence_df)) {
    return(NULL)
  }
  ts = data_scenarios_long$ts[[i]]
  incidence_df$ts = ts[incidence_df$j]
  if (data_scenarios_long$method[i] %>% str_detect("nons")) {
    incidence_df$incidence = incidence_df$incidence * 1
  }
  incidence_df = incidence_df %>%
    mutate(scenario_ID = data_scenarios_long$scenario_ID[i],
           method = data_scenarios_long$method[i],
           rep = data_scenarios_long$rep[i])
}) %>%
  bind_rows() %>%
  left_join(data_scenarios_long %>% select(-est_incidence, -ts),
            by = c("scenario_ID", "method", "rep")) %>%
  select(scenario_ID, seasonality_ratio, transmission_rates, ts, method, rep, incidence, trace)


daily_cases = lapply(valid_ix, function(i) {
  t = data_scenarios_long$ts[[i]]
  cases = data_scenarios_long$cases[[i]]
  dt = t[2] - t[1]
  tibble(t = t,
         daily_cases = cases / dt,
         scenario_ID = data_scenarios_long$scenario_ID[i],
         seasonality_ratio = data_scenarios_long$seasonality_ratio[i],
         transmission_rates = data_scenarios_long$transmission_rates[i],
         trace = data_scenarios_long$rep[i])
}) %>%
  bind_rows()

plot_cases = data_scenarios_long %>%
  unnest(c(ts, cases))

data_scenarios_daily = data_scenarios_sim %>%
  mutate(daily_cases = lapply(cases, function(x) {x/(years/annual_subdivisions)})) %>%
  unnest(c(ts, daily_cases))

ggplot(plot_data, aes(x=ts, y=incidence, color=method, group=interaction(trace, method))) +
  geom_line(alpha=0.2) +
  geom_point(data=plot_cases, aes(x=ts, y=cases, color=NULL, group=NULL), alpha=0.25) +
  # geom_line(data=plot_data, aes(y=daily_incidence, color=method, group=interaction(method, trace)), alpha=0.2) +
  facet_wrap(vars(seasonality_ratio, transmission_rates), scales="free",
             labeller = "label_both") +
  coord_cartesian(ylim = c(0, NA)) +
  scale_color_manual(values = comparison_colors)
```

What happens if we use these parameters in the full model?

```{r}
n_traces = 10
incidences = list()
for (i in seq_len(nrow(data_scenarios_long))) {
  dat = data_scenarios_long[i,]
  x = pbmclapply(seq_len(n_traces), function (j) {
    y = .simulate_cases(alpha = dat$ascertainment_rates,
                    beta = dat$radical_cure_rates,
                    lambda = sample(unlist(dat$est_lambda), 1),
                    phi_inv = 0.1,
                    delta = dat$importation_rate,
                    kappa = 1,
                    phase = 0,
                    eps = dat$seasonality_ratio,
                    N = dat$population_size,
                    index = i)
  })
  x = x["data.frame" %in% sapply(x, class)]
  
  incidences[[i]] = x %>%
    bind_rows(.id = "trace")
  
  # normalise to be per day
  dt = incidences[[i]]$ts[2] - incidences[[i]]$ts[1]
  incidences[[i]]$t = incidences[[i]]$ts - years/annual_subdivisions
  incidences[[i]]$daily_incidence = incidences[[i]]$incidence / dt
}
incidences_df = bind_rows(incidences, .id = "i") %>%
  mutate(i = as.numeric(i),
         method = data_scenarios_long$method[i],
         transmission_rates = data_scenarios_long$transmission_rates[i],
         seasonality_ratio = data_scenarios_long$seasonality_ratio[i])

# plot predictions
ggplot(incidences_df, aes(x=t/years, y=daily_incidence, color=method, group=interaction(method, trace))) +
  geom_line(alpha=0.2) +
  geom_point(data=data_scenarios_daily, aes(x=ts/years, y=daily_cases, color=NULL, group=NULL), alpha=0.25) +
  facet_wrap(vars(seasonality_ratio, transmission_rates), scales="free",
             labeller = "label_both") +
  scale_color_manual(values = comparison_colors) +
  labs("Re-simulated mean incidences using the true seasonality",
       x = "Years",
       y = "Cases/mean cases (per month)")
```

## Extension to fitting more parameters

While we previously fit only the transmission rate $\lambda$ (and overdispersion $1/\phi$ for negative-binomial models), the seasonal models introduce more complexity; here, we have parameterised seasonality by three parameters: peak/trough ratio $\epsilon$, 'sharpness' $\kappa$, and an offset from the start of the year to the to the peak `phase` (because $\phi$ is already taken). These parameters could either be implemented as equivalent to transmission suitability as a function of temperature/precipitation (as in Mordecai et al., or the MAP), or estimated simultaneously with the transmission parameters.


## Extension to include time-varying transmission levels

The Chinese and Brazilian datasets include longitudinal data for long enough to observe trends that must be due to changes in policy or environment Therefore, the seasonal Champagne model certainly will not be sufficient to explain this variation.

We will modify the Champagne model's transmission rate $\lambda$ to be time-varying, $\lambda(t)$. The functional form of $\lambda(t)$ is unclear but is constrained by the information available to fit it. For example, we will never know if a decrease in trend is due to natural decay to a low equilibrium, or due to a decreasing $\lambda(t)$. However, by finding a *reasonable* form, $\lambda(t)$ will account for long-term variation that cannot be explained by other parameters and we hope it will allow the other parameters to be recovered in scenarios where there is no form of static or dynamic steady-state at play.

At a minimum, $\lambda(t)$ will be a piecewise-constant or piecewise-linear function with breakpoints set manually when there are obvious changes in transmission intensity. This may prove sufficient. Other suggestions include particle filtering methods such as Kalman filtering.

```{r}
print("Done.")
```
_